A cortical information bottleneck during decision-making

Version 1

Abstract

Abstract

Decision-making emerges from distributed computations across multiple brain areas, but it is unclear why the brain distributes the computation. In deep learning, artificial neural networks use multiple areas (or layers) to form optimal representations of task inputs. These optimal representations are sufficient to perform the task well, but minimal so they are invariant to other irrelevant variables. We recorded single neurons and multiunits in dorsolateral prefrontal cortex (DLPFC) and dorsal premotor cortex (PMd) in monkeys during a perceptual decision-making task. We found that while DLPFC represents task-related inputs required to compute the choice, the downstream PMd contains a minimal sufficient, or optimal, representation of the choice. To identify a mechanism for how cortex may form these optimal representations, we trained a multi-area recurrent neural network (RNN) to perform the task. Remarkably, DLPFC and PMd resembling representations emerged in the early and late areas of the multi-area RNN, respectively. The DLPFC-resembling area partially orthogonalized choice information and task inputs and this choice information was preferentially propagated to downstream areas through selective alignment with inter-area connections, while remaining task information was not. Our results suggest that cortex uses multi-area computation to form minimal sufficient representations by preferential propagation of relevant information between areas.

Introduction

The brain uses multiple areas to perform cognitive functions and tasks, including decision-making, multisensory integration, attention, motor control, and timing1–10. But why distribute computation across multiple areas? One reason is that distributed computation supports important fault tolerance in the brain which allows it to compensate when dynamics in relevant areas are altered11–14. But is there also a computational benefit for distributing computations across multiple areas? In deep learning for computer vision, deeper artificial neural networks (ANNs) generally perform better and exhibit hierarchical computation, a phenomenon also observed in the visual cortex, where early layers of the neural network form representations that contain low-level details (e.g., edges) and deeper layers represent higher-level concepts (e.g., object identity) 15,16. This hierarchical computation is related to the idea of an information bottleneck: downstream areas should form representations that remove irrelevant information not necessary to do the task. For example, to know an image of a dog is a dog, we do not need to store every pixel of the image.

We unpack this more formally by first asking: what makes a representation optimal? Consider the binary classification task in Fig. 1a, where the task Y is to answer if an image, x, is a dog. The data processing inequality (DPI) states that any representation of the original image, z = f (x), where f is a function or transformation, cannot contain more information than the image itself17. Rather, a representation z will often contain less information, so that the information between random variables Z and X, denoted I(Z; X), decreases through successive transformations. This is illustrated in Fig. 1a, where different transformations of the original image decrease the mutual information I(Z; X) in the representation. But the task information contained in the representation, I(Z; Y), is similar, since all images can be used to correctly answer “is this a dog?” (Fig. 1b).

Optimal representations are formed through an information bottleneck.

(a) Consider the task Y of discerning whether the image is of a dog. Images to the right have less information than the original image (I(Z; X) is smaller) but still contain approximately the same amount of information, I(Z; Y) to perform the task: “is this a dog?” (b) The information bottleneck trades off minimality, I(Z; X) as small as possible, with sufficiency, I(Z; Y) ≈ I(X; Y). (c) Checkerboard task. The monkey reaches to the target whose color matches the checkerboard dominant color. Because there are two equally likely target configurations where the color of the left and right targets are swapped, this task unmixes the color and direction choice. (d) A minimal sufficient representation of this task is to only retain the direction decision, in this case, reach left. A cortical information bottleneck should therefore only find direction choice information in premotor and motor output areas.

The information bottleneck principle defines an optimal representation to be one that retains only the relevant or useful information for solving a task18. When a representation that is sufficient for solving the task (mathematically, I(Z; Y) ≈ I(X; Y)) is also minimal (mathematically, I(Z; X) as small as possible), machine learning theory proves these representations are the most robust to nuisances (such as the color of the dog or the background)19. In general, multi-area computation increases the minimality of a representation due to the data processing inequality: so long as these representations maintain the information to solve the task and are trainable, the deep multi-area computation produces more optimal representations for solving tasks. Although these representations contain less information than the input, they are often more useful and robust representations for solving the task19–21. Typically, these neural networks are not explicitly trained to optimize an information bottleneck objective22. However, recent studies19,20 showed that training ANNs with stochastic gradient descent implicitly minimizes an information bottleneck objective that results in minimal sufficient representations.

We hypothesized multiple areas in the brain provides a similar computational benefit by forming minimal sufficient representations of the task inputs, thus implementing a cortical information bottleneck. We tested this hypothesis by combining electrophysiological recordings in behaving monkeys, and modeling using recurrent neural networks. We recorded from the DLPFC and PMd as monkeys performed a decision-making task called the Checkerboard Task (Fig. 1c). In this task, the monkey discriminated the dominant color of a checkerboard composed of red and green squares and reached to a target matching the dominant color. Because the red and green target locations were randomly assigned to be left or right on each trial (“target configuration”), the direction decision is independent of the color decision. That is, a green color decision is equally likely to correspond to a left or right decision.

The animal’s behavioral report was either a right or left reach, determined after combining the sensory evidence with the target configuration (Fig. 1d). While color is initially needed to solve the task, the minimal sufficient representation of the task to generate the correct output is a representation of only the direction decision without the color decision or the target configuration. A cortical information bottleneck would therefore predict that upstream areas should contain information about the task inputs and decision-making process, including the target configuration, perceived dominant color of the checkerboard, and direction choice, while downstream areas should only contain the direction choice (in Fig. 1d, “reach left”).

Consistent with the predictions of the information bottleneck principle, we found that DLPFC has information about the color, target configuration, and direction. In contrast, PMd had little to no information about target configuration and color, but strongly represented the direction choice. PMd therefore had a minimal and sufficient representation of direction. We then trained a multi-area RNN to perform this task. We found that the RNN faithfully reproduced DLPFC and PMd activity, enabling us to propose a mechanism for how cortex uses multiple areas to compute a minimal sufficient representation.

Results

The checkerboard task involves multiple brain areas

We trained monkeys to discriminate the dominant color of a central static checkerboard (15 × 15 grid) composed of red and green squares (Fig. 1c). The number of red and green squares was random on each trial, leading to different levels of discrimination difficulty. The signed color coherence linearly indicates the degree to which color is dominant in the checkerboard, with − 1 corresponding to completely green, 0 to equal numbers of red and green squares, and +1 to completely red. If the monkey reached to the target matching the dominant checkerboard color, the trial was counted as a success and the monkey received a juice reward. Critically, the color decision (red or green) was decoupled from the direction decision (left or right) because the left and right target identities were random on every trial. Fig. 2a,b shows the monkey’s psychometric and reaction time (RT) behavior on this task. Monkeys made more errors and reacted more slowly for more ambiguous checkerboards compared to the almost completely red or completely green checkerboards23.

DLPFC and PMd recordings during the checkerboard task.

(a) Psychometric and (b) reaction time curves of the monkey. X-axes in both (a) and (b) depict signed coherence which indicates the relative amount of red vs. green in the checkerboard. (c) Example DLPFC and (d) PMd PSTHs aligned to checkerboard onset. Red and green traces correspond to red and green color choices, respectively. Dotted and solid traces correspond to right and left direction choices, respectively. Data are smoothed with a 25 ms Gaussian and averaged across trials. (e) PCs 1, 3, and 4 for DLPFC and (f) PCs 1,2,3 for PMd. (g) Results of dPCA analysis for DLPFC and (h) PMd showing the dPCs for direction, target configuration, and color. (i) Histogram (across sessions) of direction, color, and target configuration decode accuracy and (j) usable information for DLPFC and PMd. The large variance in recordings is due to across-session variance.

We used linear multi-contact electrodes (U and V-probes) to record from the DLPFC (2819 single neurons and multiunits) and PMd (996 single neurons and multi units) as the monkeys performed these tasks. Sample peri-stimulus time histograms (PSTHs) for neurons in DLPFC and PMd are shown in Fig. 2c, d, respectively, where solid (dotted) lines correspond to left (right) reaches and color (red or green) denotes the color decision. DLPFC PSTHs in Fig. 2c separate based on direction choice, target configuration, and color choice, whereas PMd PSTHs primarily separate based on the direction choice, and only very modestly with target configuration or color.

Together, these examples demonstrate that DLPFC and PMd single units exhibit activity reflecting the decision-making process, implicating multiple brain areas in decision-making. Further, DLPFC likely contains multiple task-relevant signals signals, whereas PMd contains only direction choice related signals necessary for the behavioral report in the task. In the next sections, we use dimensionality reduction, decoding, and information theory to quantify the extent of color, target configuration, and direction representations in DLPFC and PMd at the population level and show that these physiological observations are consistent with the information bottleneck principle. We then use recurrent neural network models to build a mechanistic hypothesis for how an information bottleneck could be implemented.

Evidence for a cortical information bottleneck between DLPFC and PMd

Our single neuron examples suggest that neuronal responses in DLPFC are modulated by color choice and target configuration, but PMd neurons generally are not. Our hypothesis is that these cortical representations in DLPFC and PMd are consistent with the information bottleneck principle. The direct prediction of this hypothesis is that the PMd population activity should contain a minimal and sufficient representation of the behaviorally relevant output – the direction choice – while upstream DLPFC population activity should represent multiple task-relevant variables.

To study this at the population level, we performed principal components analysis (PCA) on DLPFC and PMd neural population activity. In these PCA trajectories, we subtracted the condition-independent component of the signal to better highlight representations of direction, target configuration, and color. DLPFC and PMd exhibited qualitatively different neural population trajectories (Fig. 2e,f). First, after target onset, DLPFC trajectories separated as a function of the two target configurations, and at the time of checkerboard onset (purple dots), DLPFC activity further separated into four distinct trajectories based on the four possible color × direction outcomes (green left, green right, red left, red right, Fig. 2e). Thus, DLPFC contains information about target configuration, color choice, and direction choice. Note, we chose PC 1, 2 and 4 of DLPFC as they provided a better visualization of target configuration signal between target and checkerboard. In contrast, PMd trajectories in Fig. 2f did not exhibit target-configuration-specific steady state responses. Thus, at the point of checkerboard onset (purple dots), trajectories overlapped in the top 3 principal components, and only separated based on the direction, but not color, choice.

To quantify these these differences, we performed demixed principal component analysis (dPCA) on the DLPFC and PMd population activity (Fig. 2g,h). DLPFC and PMd activity both exhibited strong condition independent activity (82% and 86% variance, respectively). DLPFC activity represented the target configuration, but PMd did not (Fig. 2g,h, target configuration dPC). dPCA also identified principal axes that maximized variance related to the direction choice, color choice, and target configuration. In DLPFC, the top direction choice, color choice, and target configuration axes captured 7.1%, 1.5%, and 0.9% of the population activity. In PMd, these values were 10.6%, 0.2%, and 0.2%. Across all direction, color, and target configuration axes, the dPCA variance captured for DLPFC was 11%, 3%, 5%, while for PMd it was 12%, 1%, 1%. This dPCA analysis provides further evidence that DLPFC represents direction, target configuration, and color while PMd has nearly minimal representations of color and target configuration, consistent with the information bottleneck principle.

Our dPCA results with trial-averaged firing rates suggest that axes associated with color and target configuration in PMd have very little variance associated with them. However, these results do not rule out the possibility that there is decodable information about these task-related variables on single trials. We performed two other analyses to assess if there is an information bottleneck and that PMd contains a minimal sufficient representation. We calculated the decode accuracy and an estimate of mutual information for direction, color, and target configuration in DLPFC and PMd population activity. We decoded direction, color, and target configuration from DLPFC and PMd sessions where we recorded a small population of neurons using a support vector machine (see Methods). To estimate mutual information, we quantified the Usable Information, a variational lower bound to mutual information that can be computed on high-dimensional data through estimating cross-entropy loss20,21 (see Methods).

We found that for many DLPFC sessions we could reliably decode direction, color, and target configuration well above chance (histogram in Fig. 2i, mean decoding accuracy across sessions: direction 85%, target configuration 59%, color 57%, for details on decoding, see Methods). To conservatively assess significance, we shuffled the trials for each session 100 times and obtained a surrogate decoding accuracy for each session, and then tested if the true accuracy was greater than the 99th percentile of the shuffled accuracy distribution (Wilcoxon Sign Rank test: p < 0.005 for target configuration, color, and direction choice). In DLPFC, we could therefore significantly decode direction choice, target configuration, and color choice. In contrast in PMd, we rarely found sessions where target configuration and color could be reliably decoded well above chance (mean accuracy across sessions: direction 87%, target configuration 52%, color 51%), as shown in Fig. 2i. Again, we compared the true decode accuracy to a surrogate decoding accuracy obtained by shuffling the trials. We found that only direction could be decoded well above chance (Wilcoxon sign-rank test: direction choice p < 0.005, p = 0.99 for both target configuration and color choice). The differences in mean decoding accuracy in DLPFC and PMd were significant for only color and target configuration (p < 0.001, ranksum test), but not direction (p = 0.208, ranksum test). Thus, both DLPFC and PMd contain signifcantly more direction information but DLPFC contains more color choice and target configuration information consistent with the information bottleneck principle.

We also quantified usable information for DLPFC and PMd (Fig. 2j). For this analysis, we restricted it to sessions with significant decode accuracy with a session considered to have a significant decodability for a variable if the true accuracy was above the 99th percentile of the shuffled accuracy for a session. DLPFC had sessions with non-zero usable information for direction, color, and target configuration (average direction information: 0.56 bits, target configuration: 0.07 bits, color: 0.06 bits; 100%, 78%, 53% sessions above 99 percentile of shuffled accuracy, respectively) while PMd only had non-zero usable information for direction (average direction information: 0.64 bits, target configuration: 0.013 bits, color: 0.005-bits; 98%, 26%, 20% sessions above 99 percentile of shuffled accuracy, respectively). The differences in usable information in DLPFC and PMd were also significant for only color and target configuration (p < 0.001, ranksum test), but not direction (p = 0.107, ranksum test). Together, these results indicate that PMd had a more minimal representation of task inputs, particularly color and target configuration, than DLPFC.

Our dPCA and decoding results are consistent with the existence of a cortical information bottleneck between DLPFC and PMd that reduces the amount of target configuration and color information in PMd while preserving the direction choice information necessary to solve the task. We next modeled this multi-area information bottleneck to develop a mechanistic hypothesis for how this cortical information bottleneck could be computationally implemented.

A multi-area recurrent neural network model of DLPFC and PMd

To develop a mechanistic hypothesis for this cortical information bottleneck, we studied our previously reported multi-area RNN to perform the Checkerboard task (Fig. 3a)24. We chose to use this multi-area RNN because prior work demonstrated this RNN, like our PMd data, has a minimal color representation in Area 3. The RNN had 3 areas, obeyed Dale’s law25, and had approximately 10% feedforward and 5% feedback connections between areas based on projections between prefrontal and premotor cortex in a macaque atlas26. RNN psychometric and RT curves for the multi-area RNN exhibited similar behavior to monkeys performing this task (Fig. 3b,c; across several RNNs, see Fig. S1).

RNN modeling of the CB task.

(a) Multi-area RNN configuration. The RNN received 4 inputs. The first two inputs indicated the identity of the left and right targets, which was red or green. These inputs were noiseless. The last two inputs indicated the value of the signed color coherence (proportional to amount of red in checkerboard) and negative signed color coherence (proportional to amount of green in checkerboard). We added independent Gaussian noise to these signals (see Methods). The network outputted two analog decision variables indicating evidence towards the right target (solid line) or left target (dashed line). A decision was made in the direction of whichever decision variable passed a preset threshold (0.6) first. The time at which the decision variable passed the threshold was defined to be the reaction time. (b,c) Psychometric and reaction time curves for exemplar multi-area RNN. (d) Area 1 and Area 3 principal components for exemplar RNN. (e) CCA correlation between each area and DLPFC principal components (left) and PMd principal components (right). DLPFC activity most strongly resembles Area 1, while PMd activity most strongly resembles Area 3. See also Fig. S3 where we computed CCA as a function of the number of dimensions. (f) Relative dPCA variance captured by the direction, color, and target configuration axes. Normalization makes direction variance equal to 1. Area 1 (3) variances more closely resemble DLPFC (PMd). (g) Area 1 has significantly higher decoding accuracies and (h) usable information compared to Area 3, consistent with DLPFC and PMd.

Remarkably, even though the multi-area RNN was in no way regularized to reproduce DLPFC and PMd activity, activity in Area 1 resembled neural responses in DLPFC and Area 3 resembled PMd (Fig. 3d,e). Like DLPFC, Area 1 had four distinct trajectories corresponding to the four possible task outcomes and represented target configuration, direction choice, and color choice (Fig. 3d and see Fig. S2). In contrast, Area 3 population trajectories primarily separated based on direction and not by target configuration or color — remarkably similar to the trajectories observed in PMd.

We performed CCA to assess the similarity between the empirical neural trajectories to each RNN area’s neural trajectories (see Methods). The CCA analysis suggested that Area 1 exhibited the strongest resemblance to DLPFC, while Area 3 most strongly resembled PMd activity (Fig. 3e). These results show that a multi-area RNN reproduced similar behavior to the monkey, and further that it did so with architecturally and qualitatively distinct areas that strongly resembled the physically distinct DLPFC and PMd cortical areas.

The RNN activity differs from cortical activity in two important ways. First, RNNs generally had a significantly smaller variance condition-independent signal (46.7% and 49.4% average variance in Area 1, and 3, respectively) than in DLPFC and PMd (82% and 86% variance, respectively). One possible explanation is that condition-independent variance in PMd is associated with a trigger signal, likely from the thalamus27, and these RNNs do not output arm kinematics, forces, or electromyography (EMG).

Similarly, in DLPFC, we did not explicitly model the target and checkerboard inputs to have large onset signals that are often associated with visual stimulation. This significant condition independent variance in the neurophysiological data may therefore make decoding more difficult since there is relatively lower variance representing direction, color, or target configuration. While our CCA analysis was performed with the condition-independent signal removed, this difference impacts both dPCA and decoding results. In general, we found that RNN exhibited trends observed in the neurophysiological data more strongly, including more variance captured for direction, color, and target configuration, as well as higher decoding accuracies. We therefore compared the relative, rather than absolute, trends in RNN activity and DLPFC for dPCA and decoding analyses for the purposes of identifying a RNN information bottleneck similar to the cortical information bottleneck.

We found that the 3-area RNN exhibited similar trends to DLPFC and PMd activity in dPCA variance and decoding accuracy. When comparing only the top axis for direction, color, and target configuration, DLPFC activity had relatively large variance captured along the direction axis (7.1% variance captured), followed by relatively weaker representations for target configuration (1.5%) and color (0.9%). Area 1 activity had similar relative trends, with the direction axis explaining 30.9% variance, followed by target configuration (13.3%) and color (5.6%). In Fig. 3f, we show the similarity in relative trends in DLPFC and Area 1 by normalizing these quantities by the variance captured by the direction axis. Meanwhile, PMd activity exhibited more direction-related variance than DLPFC (10.6% variance captured) and did not exhibit a significant target configuration and color axis representation (0.2% for both axes).

Likewise, Area 3 had a stronger representation of direction (48.5% variance captured) than in Area 1, but negligible target configuration and color axes variance (0.1% for both axes), demonstrating the same relative trend (Fig. 3f). These results show that Area 1 more strongly resembles DLPFC and Area 3 more strongly resembles PMd in relative dPCA variance.

We next evaluated the decode accuracy and usable information in the multi-area RNN. We found Area 1, like DLPFC, had significant information for direction, target configuration, and color. Direction, color, and target configuration could be decoded from Area 1 population activity at accuracies of 94.4%, 93.4%, and 99.0%, respectively, corresponding to 0.81, 0.79, and 0.92 bits of usable information (Fig. 3g,h). In contrast, Area 3 direction decode accuracy was 99.4%, while color and target configuration accuracy were significantly lower (51.1% and 54.3%, respectively). This corresponded to 0.97, 0.0023, and 0.0078 bits of usable information for direction, target configuration, and color, respectively.

Together, these results show that our multi-area RNN exhibited distinct areas that resembled DLPFC and PMd activity, and also implemented an information bottleneck so that its output area only had primarily direction information and less color and target configuration information. This multi-area RNN therefore implements a candidate mechanism for how a cortical information bottleneck could be implemented between DLPFC and PMd.

Mechanistic features of the DLPFC and PMd bottleneck: partial orthogonalization and selective propagation

The multi-area RNN contains representations consistent with the information bottleneck principle — its input and output areas resemble DLPFC and PMd. The RNN therefore models many aspects of our physiological data and is therefore a candidate system to understand how multiple areas could lead to the empirically observed minimal sufficient representations. The unique advantage of the RNN is that we know the firing rates in each area as well as the within and inter-areal connections, which allows us to investigate how the multi-area RNN deemphasizes color information through an IB. We reasoned that such an information bottleneck could be implemented in three ways: Color information may be (1) primarily attenuated through recurrent neural dynamics, (2) primarily attenuated through inter-areal connections, or (3) attenuated through a combination of recurrent dynamics and inter-areal connections. This is illustrated in Fig. 4a.

IB hypotheses and mechanism.

(a) Candidate mechanisms for IB. (b) Axes overlap of the direction, color, and target configuration axes for DLPFC and RNN data. The direction axis is more orthogonal to the color and target configuration axes. (c) Projections onto the potent space of the intra-areal dynamics for each area. We computed the potent projection of the direction axis, color axis, and a random vector with each area’s intra-areal dynamics matrix. We found intra-areal dynamics amplify color information in Area 1, and do not selectively attenuate color information in Areas 2 and 3. (d) Illustration depicting how the orientation of the axes affect information propagation. Information on the direction axis (orange) can be selectively propagated through inter-areal connections which information on the color axis (maroon) is not. (e) Inter-areal hypotheses. (f) Projections onto the potent space between areas for the color axis, direction axis, and random vector. Regardless of the dimension of the potent space, the direction axis is preferentially aligned with the potent space, indicating the information along this axis propagates, while the color axis is approximately randomly aligned. We emphasize the high alignment of the direction axis: the direction axis has a stronger alignment onto the first potent dimension of W21 than the remaining dimensions combined. Meanwhile, the color axis is aligned at nearly chance levels, and will therefore be propagated significantly less than the direction axis. Shading indicates s.e.m.

To test these hypotheses, we first quantified how color, target configuration, and direction information was represented in the animals and our network. We performed dPCA in the different areas to identify demixed principal components that represented the corresponding information. In DLPFC, we quantified the overlap of the dPCA principal axes for target configuration, color, and direction. While the target configuration and color axes were relatively aligned (dot product, DP: 0.52), the direction axis was closer to orthogonal to the color axis (DP: 0.18) and the target configuration axis (DP: 0.35), as shown in Fig. 4b. These results suggest that DLPFC partially orthogonalizes information about the direction choice from the color choice and target configuration. We also observed these trends in the RNN, albeit more strongly. In DLPFC-resembling Area 1, we observed the target configuration and color were also highly aligned (DP: 0.95) but that the direction axis was more orthogonal to the color axis (DP: 0.13) and the target configuration axis (DP: 0.12). In our simulations, the reported values reflect the mean across 8 networks trained with the same hyperparameters. A candidate mechanism for this orthogonalization, found by performing dynamical analyses on the RNN, is shown in Fig. S5.

The advantage of our model is that both the intra-areal dynamics and inter-areal connectivity matrices are known. We analyzed how these axes were aligned with the intra-areal recurrent dynamics and inter-areal connectivity matrices to identify which hypothesis explained how the RNN implemented the information bottleneck. To do so, we performed singular value decomposition (SVD) on these matrices. We defined a k-dimensional “potent space” to be the right singular vectors corresponding to the k largest singular values of the matrix. The “null space” is the orthogonal complement of the potent space, which comprises the remaining d − k smallest singular vectors, where d refers to number of columns of the matrix. The null projection magnitudes are equal to one minus the potent projection. We quantified how the color and direction axis were aligned with these potent and null spaces (see Methods). This enabled us to study if the emergence of minimal sufficient representations was due to: (1) relative amplification of the direction information with respect to a random vector, (2) relative suppression of the color/target configuration information with respect to a random vector, or (3) a combination of both. Finally, we focused our analyses on Area 1 recurrent dynamics and the inter-areal connections between Areas 1 and 2 (W21) because color information is significantly attenuated by Area 2 (dPCA color variance in Area 2: 0.14%). The same analyses applied to downstream areas are shown in Fig. S8.

We first tested the hypothesis that the RNN IB is implemented primarily by recurrent dynamics (left side of Fig. 4a). We quantified how the color and direction axis were aligned with these potent and null spaces of the intra-areal recurrent dynamics matrix of Area . The axes found through dPCA are proxies for how task information is represented in the neural activity over the trial, while the dynamical activity during a trial depends on the complex interaction between left and right singular vectors of the recurrent matrix (and task inputs)28. We use the alignment of the (fixed) dPCA axes with the right singular vector as a proxy for the amount of information propagated through recurrence. In Area 1, we found significant alignment of the color axis with the top right singular vectors (potent space) of the recurrent dynamics matrix (Fig. 4c). Additionally, we also performed an alternative analysis where we compared input and activity representations of color discriminability and direction discriminability for our exemplar network. We observe an amplification, not a reduction, in color discriminability with respect to the inputs in Area 1 (Fig. S6) consistent with the amplification observed in Fig. 4c. These findings argues against the hypothesis that recurrent dynamics preferentially attenuate color information by projecting it into a nullspace of the recurrent dynamics. Rather, these data suggest that Area 1 has significant color information in its potent space, indicating that the recurrent computation amplifies color information. In Areas 2 and 3, the color axis (which had small variance of 0.14% and 0.07% in Areas 2 and 3, respectively) was again typically more strongly aligned with than a random vector, (Fig. S8). In summary, the dPCA and discriminability analyses suggest that the network did not use recurrent dynamics to attenuate color information, and is therefore inconsistent with the hypothesis that the IB is primarily implemented through intra-areal recurrent dynamics.

Our alternative hypothesis is that color information is primarily attenuated through inter-areal connections. This is schematized in Fig. 4d, where inter-areal connections propagate activity along the Area 1 direction axis (orange) to Area 2, but attenuate Area 1 color axis activity (maroon). To test this hypothesis, we quantified how the color and direction axis were aligned with these potent and null spaces of the inter-areal matrices. This enabled us to quantify the alignment of the direction and color axes with the inter-areal potent and null spaces and specifically determine how direction and color information were differentially propagated (Fig. 4e). Inter areal connections could attenuate color information by aligning the color axis with the null space of W21 (Hypothesis 1 in Fig. 4e), propagate information preferentially (Hypothesis 2 in Fig. 4e), or both attenuate and propagate information (Hypothesis 3 in Fig. 4e).

We calculated the projections for both the color and choice axes on to the potent space for theconnection matrix from area 1 to area 2 (W21) The projections onto the potent space are shown in Fig. 4f for the color and direction axis. We found the direction axis was more aligned with the potent space and the color axis was more aligned with the null space. In fact, the direction axis was consistently most aligned with the top singular vector of the W21 matrix, on average more than the remaining d − 1 singular vectors. In contrast, the color axis was aligned to a random vector. This suggests that learning in the multi-area recurrent network involved aligning the relevant information (in the activations) with the top singular vector (governed by the learned parameters of the feedforward matrix). These results indicate that direction information is preferentially propagated to subsequent areas, while color information is aligned with a random vector thus maximally consistent with the “propagate” hypothesis shown in Fig. 4e.

Such alignment of the direction axis with the top singular vector of the connection matrix isn’t trivial: the potent space depends on the parameter W21 learned during training, while the direction axis is not a parameter but a dPCA axis computed from Area 1 activity. This alignment was robust to the dimension of the effective potent space, and was consistent across networks with varying feedforward connectivity percentages (10%, 20%, 30%, 50%, 100%). Further, we found that W21 in unconstrained 3 area networks had significantly reduced alignment of the direction axis with the top singular vectors (Fig. S8d).

In summary, the multi-area RNN information bottleneck is primarily implemented through preferential propagation of direction information through inter-areal connections. Recurrent dynamics play a role in processing color information and target configuration to arrive at direction choice information. In Area 1, RNN dynamics actually amplify color information. Our results are therefore most consistent with the hypothesis that the IB is implemented primarily through inter-areal connections, not recurrent dynamics, in Fig. 4a.

Effect of network architecture and training hyperparameters on the information bottleneck

We next assessed the network architectures and hyperparameters that influenced the formation of minimal sufficient representations during the Checkerboard task. We swept RNN architectural parameters and machine learning hyperparameters to assess what variables were important for learning minimal sufficient representations without color information. Specifically, we varied the connectivity type (unconstrained connectivity vs Dale’s law and varying proportion of connections), the percentage of unconstrained feedforward connections, the percentage of feedforward E to I connections, the percentage of E to E connections, the number of areas (from 1 to 4), the number of artificial networks, L2 weight regularization, L2 rate regularization, and the learning rate. In our sweeps we quantified the color (and direction) variance and accuracy in the last area.

We generally observed minimal sufficient representations in the last area so long as there was a sufficient connection bottleneck between RNN areas. In unconstrained networks, shown in Fig. 5a, color variance and decode accuracy decreased as the percentage of feedforward connections between areas decreased, though the representations were not minimal. We incorporated Dale’s law with 80% E and 20% I neurons following Song et al.25 into subsequent sweeps (Fig. 5b-e). Minimal representations with chance color decode accuracy emerged when the percentage of feedforward E to I connections was 2 − 5% or less (the overall percentage of feedforward E to E was fixed at 10% following a macaque atlas). We also found that when there was no feedforward inhibition, but when we varied the percentage of feedforward or feedback E-to-E connections RNNs generally had nearly minimal representations (Fig. 5c, and Supp. Fig. S9). We observed that as long as there were 3 or 4 areas, there was a large decrease in color information in the last area (Fig. 5d) quantified by decoding, though note that there was a large drop in color variance for 2 area networks. These results suggest that multi-area networks, with a feedforward connection bottleneck tend to produce more minimal representations for the Checkerboard task.

Robustness of the information bottleneck across hyperparameters and computational advantage.

Varying (a) proportion of feedforward connections in an unconstrained network, (b) E-I connections in a Dale’s law network, (c) the proportion of feedforward and feedback E-E connections in a Dale’s law network, (d) the number of areas, and (e) the machine learning hyperparameters revealed that Area 3 color variance and color accuracy decrease as long as there is a connectivity bottleneck between areas. (f) Summary of these results quantifying usable information.

We also varied machine learning hyperparameters (Fig. 5e) to assess the extent to which the information bottleneck was present. To prevent an exponential search space, we fixed the architecture to the exemplar network used in this study and tested one hyperparameter at a time. We varied the number of artificial units in the network, the L2 weight regularization, the L2 rate regularization, and the learning rate. At each hyperparameter setting, we trained a total of 8 multi-area RNNs. Our exemplar network consistently exhibited little to no Area 3 color information across every hyperparameter setting we chose, suggesting that the presence of the information bottleneck is not a result of a particular choice of machine learning hyperparameters.

We again summarized all sweeps by calculating the “Usable Information”20 to quantify the direction and color information in RNNs, as shown in Fig. 5f and the results reaffirmed conclusions from the variance and decoding analyses. Together, these results suggest that a connection bottleneck in the form of neurophysiological architecture constraints was the key design choice leading to RNNs with minimal color representations and consistent with the information bottleneck principle.

Discussion

The goal of this study was to investigate if predictions from the information bottleneck principle in machine learning and information theory are also observed in cortical circuits. The information bottleneck principle defines an optimal representation to be one that retains only the relevant or useful information for solving a task18. This principle has been applied to explain the success of deep networks19,29, by forming minimal sufficient representations of task inputs, leading to better generalization bounds and invariance to noise19. We explored whether such a principle could explain cortical representations across different areas during a visual perceptual decision making task. We found that later areas of cortex along a sensorimotor transformation (in PMd) only represented the behavioral report, that is the action choice, while earlier areas had stronger input representations and performed relevant computations to define the behavioral report (in DLPFC). To better understand how such a phenomenon could be implemented in cortex, we trained many artificial multi-area RNNs to perform this task. Surprisingly, we also observed that RNNs formed minimal sufficient representations across a range of hyperparameter settings, suggesting the formation of minimal sufficient representations may be a more general feature of multi-area computation.

Given the “full-observability” of our multi-area models, we were able to analyze the learned weight matrices and understand how the network converged to transform task inputs into minimal sufficient representations by the output area. In particular, we found that the output-relevant direction information was preferentially propagated between areas by having the largest overlap with the top singular vector of the learned feedforward matrices. In contrast, color information was almost randomly propagated through feedforward connections. This mechanism is related to prior work on output potent and output null subspaces30 and communication subspaces1, with the important difference that color information isn’t preferentially projected to a nullspace, but is aligned similarly to any random vector. Preferential alignment with a cortical nullspace is therefore not necessary to achieve an IB — color information may be attenuated through random alignment to the communication subspace. This solution (random alignment) poses less constraints on inter-areal connectivity than a solution that preferentially propagates direction information while also preferentially projecting color information to a nullspace.

Our results are also consistent with recent work proposing that cortical areas convey information through communication subspaces. One observation in communication subspaces is that they do not merely propagate the directions of highest variance1. We also observed this phenomenon for the W21 connectivity matrix, which communicates information from Area 1 to Area 2. Color activity had significant variance in Area 1 (see Fig. S5). Inter-areal connections must therefore not merely propagate the highest variance dimensions of a preceding area, otherwise color information would be conveyed to Area 2. Consistent with this, we found that while the top 2 PCs capture 97.7% excitatory unit variance, the top 2 readout dimensions of W21 only captured 40.0% of Area 1’s excitatory unit neural variance (Fig. S7). Hence, inter-area connections are not aligned with the most variable dimensions, but are rather aligned to preferentially propagate certain types of information — a result consistent with a recent study analyzing links between activity in V1 and V21.

We find minimal sufficient representations in PMd and in the later areas of our recurrent network models. Do such representations have any advantages? One possibility is that in cortex a minimal sufficient representation provides energetic benefits31,32. Another possibility is that such a representation provides a computational advantage. This is an open question that is still somewhat unresolved in the machine learning community, with representation learning approaches that maximize mutual information between representations and inputs also leading to useful task representations33, in addition to compressed representations19,20. The information contained in the representation of a neural network is related to the “Information in the Weights”34, which can be quantified using the Fisher Information35–37, a measure of sensitivity to perturbations. This “Information in the Weights” view would predict that minimal sufficient representations have smaller Fisher information and are therefore less sensitive to (local) perturbations in the readout weights. In the context of deep networks, it has been proposed that minimal sufficient representations simplify the role of the output readout or classifier19. Further, a minimal sufficient representation with respect to a family of probabilistic decoders/classifiers will provably generalize better38.

Although finding a resolution to this debate in machine learning is beyond the scope of this paper, we assessed if minimal RNNs exhibited any qualities consistent with machine learning predictions. We explored whether minimal sufficient representations would simplify the readout, which we quantified by measuring the model’s performance in response to perturbations to the readout weights. We found that 3-area networks with minimal color information (particularly networks in Fig. 5b with no feedforward E-to-I connectivity) were less sensitive to perturbations than corresponding networks with significant color information (networks in Fig. 5a with 10% unconstrained feedforward connectivity, see Fig. S11). We also found that these networks differed significantly in readout complexity, with 3-area networks with minimal color information exhibiting simpler and sparser readouts (Fig. S11). However, we did not observe a clear trend between perturbation sensitivity and usable color information across random initializations (Fig. S11) for a fixed parameter setting (networks with 10% feedforward inhibition in Fig. 5b). An interesting venue for future work is to further examine the potential advantages of a minimal sufficient representation. Such findings would be valuable to the machine learning and neuroscience community. In our study, several factors including recurrent connectivity, multiple areas, and E/I populations make theoretical study of this question difficult. It is likely that studying this question requires simplifying the setting. For example, it likely makes sense to first focus on feedforward networks with a variable amount of task input information, similar to the generalized checkerboard-task used in20.

Our task could be solved with or without feedback connections with equivalent performance, indicating that feedback was not necessary to solve the task (Fig. S9). Minimal sufficient representations were found in both purely feedforward RNNs or RNNs with feedback (Fig. S9). When the model had feedback connections, we observed that feedback connections between Areas 2 and 1 preferentially conveyed direction information. Due to the presence of choice related signals in several cortical areas, these feedback connections may also play a role in computation of the direction choice. Another perspective on feedback signals is that they may related to error signals used for learning39. Multi-area networks may help understand and develop new hypotheses for physiological studies of feedforward and feedback computation40,41, and more generally distributed processing for decision-making and cognition8,42.Future research may use carefully designed tasks in conjunction with multi-area RNNs to better understand the role of feedback in computation.

The multi-area RNN also provides several testable hypotheses. First, because the information bottleneck simplifies the readout, it suggests a simple readout from “output” areas of cortex (e.g PMd or Motor Cortex). In our multi-area RNN, we found that the output area was comprised of pools of neurons that represent left or right reaches43, enabling a simple readout (Fig. S10b, c.). In particular, the connectivity matrix for our area 3 was composed of two pools of excitatory neurons with a common inhibitory pool. Overall, this connectivity and winner-take-all architecture is consistent with our PMd activity, which appears to implement winner-take-all dynamics between a pool of neurons representing left and right reaches (Fig. S10h,i.) Second, due to selective propagation of inter-areal information, direction axis activity in DLPFC should be more predictive of activity in downstream regions such as PMdr and PMd than activity in the top PCs. Simultaneous recordings of DLPFC and PMd would help test this hypothesis.

Materials and Methods

Task and training details

Somatomotor reaction time visual discrimination task and recordings from PMd

The task, training and electrophysiological methods used to collect the data used here have been described previously23 and are reviewed briefly below. All surgical and animal care procedures were performed in accordance with National Institutes of Health guidelines and were approved by the Stanford University Institutional Animal Care and Use Committee and the Boston University Institutional Animal Care and Use Committees.

Two trained monkeys (Ti and Ol) performed a visual reaction time discrimination task. The monkeys were trained to discriminate the dominant color in a central static checkerboard composed of red and green squares and report their decision with an arm movement. If the monkey correctly reached to and touched the target that matched the dominant color in the checkerboard, they were rewarded with a drop of juice. This task is a reaction time task, so that monkeys initiated their action as soon as they felt they had sufficient evidence to make a decision. On a trial-by-trial basis, we varied the signed color coherence of the checkerboard, defined as (R − G)/(R + G), where R is the number of red squares and G the number of green squares. The color coherence value for each trial was chosen uniformly at random from 14 different values arranged symmetrically from 90% red to 90% green. Reach targets were located to the left and right of the checkerboard. The target configuration (left red, right green; or left green, right red) was randomly selected on each trial. Both monkeys demonstrated qualitatively similar psychometric and reaction-time behavior. 996 units were selected from PMd of Ti (n=546) and Ol (n=450) and 2819 units were recorded from DLPFC of Ti while they performed the task23. Monkey Ol and Ti’s PMd units both had low choice color probability. Reported analyses from PMd data use units pooled across Monkey Ol and Ti.

RNN description and training

We trained a continuous-time RNN to perform the checkerboard task. The RNN is composed of N artificial neurons (or units) that receive input from Nin time-varying inputs u(t) and produce Nout time-varying outputs z(t). The RNN defines a network state, denoted by x(t) ∈ ℝN ; the ith element of x(t) is a scalar describing the “currents” of the ith artificial neuron. The network state is transformed into the artificial neuron firing rates (or network rates) through the transformation: where f (·) is an activation function applied elementwise to x(t). The activation function is typically nonlinear, endowing the RNN with nonlinear dynamics and expressive modeling capacity44. In this work, we use f (x) = max(x, 0), also known as the rectified linear unit, i.e., f (x) = relu(x). In the absence of noise, the continuous time RNN is described by the equation where τ is a time-constant of the network, Wrec ∈ ℝN ×N defines how the artificial neurons are recurrently connected, brec ∈ ℝN defines a constant bias, maps the RNN’s inputs onto each artificial neuron, and ϵt is the recurrent noise. The output of the network is given by a linear readout of the network rates, i.e., where maps the network rates onto the network outputs.

We trained RNNs to perform the checkerboard task as follows. For all networks, unless we explicitly varied the amount of units, we used Nin = 4, N = 300, and Nout = 2.

The four inputs were defined as:

Whether the left target is red (−1) or green (+1).

Whether the right target is red (−1) or green (+1).

Signed coherence of red (ranging from -1 to 1), (R − G)/(R + G).

Signed coherence of green (ranging from -1 to 1), (G − R)/(R + G). Note that, prior to the addition of noise, the sum of the signed coherence of red and green is zero.

The inputs, u(t) ∈ ℝ4, were defined at each time step, t, in distinct epochs. In the ‘Center Hold’ epoch, which lasted for a time drawn from distribution 𝒩 (200 ms, 502 ms2), all inputs were set to zero. Subsequently, during the ‘Targets’ epoch, which lasted for a time drawn from distribution 𝒰 [600 ms, 1000 ms], the colors of the left and right target were input to the network. These inputs were noiseless, as illustrated in Fig. 3a, to reflect that target information is typically unambiguous in our experiment. Following the ‘Targets’ epoch, the signed red and green coherences were input into the network during the ‘Decision’ epoch. This epoch lasted for 1500 ms. We added zero mean independent Gaussian noise to these inputs, with standard deviation equal to 5% of the range of the input, i.e., the noise was drawn from 𝒩 (0, 0.12). At every time point, we drew independent noise samples and added the noise to the signed red and green coherence inputs. We added recurrent noise Et, adding noise to each recurrent unit at every time point, from a distribution 𝒩 (0, 0.052). Following the ‘Decision’ epoch, there was a ‘Stimulus Off’ epoch, where the inputs were all turned to 0.

The two outputs, z(t) ∈ ℝ2 were defined as:

Decision variable for a left reach.

Decision variable for a right reach.

We defined a desired output, zdes(t), which was 0 in the ‘Center Hold’ and ‘Targets’ epochs. During the ‘Decision’ epoch, zdes(t) = 1. In the ‘Stimulus Off’ epoch, zdes(t) = 0. In RNN training, we penalized output reconstruction using a mean-squared error loss, The set 𝒯 included all times from all epochs except for the first 200 ms of the ‘Decision’ epoch from the loss. We excluded this time to avoid penalizing the output for not immediately changing its value (i.e., stepping from 0 to 1) in the ‘Decision’ epoch. Decision variables are believed to reflect a gradual process consistent with non-instantaneous integration of evidence, e.g., as in drift-diffusion style models, rather than one that steps immediately to a given output. To train the RNN, we minimized the loss function: where

‖A‖F denotes the Frobenius norm of matrix A

λin = λrec = λout = 1, λr = 0 to penalize larger weights.

λΩ = 2

ℒΩ is a regularization term that ameliorates vanishing gradients proposed and is described in prior literature25,45.

During the training process, we also incorporated gradient clipping to prevent exploding gradients45. Training was performed using stochastic gradient descent, with gradients calculated using backpropagation through time. For gradient descent, we used the Adam optimizer, which is a first order optimizer incorporating adaptive gradients and momentum46.

Every 200 or 500 training epochs, we generated 2800 cross-validation trials, 100 for each of the 28 possible conditions (14 coherences × 2 target configurations). For each trial, there was a correct response (left or right) based on the target configuration and checkerboard coherence. When training, we defined a “correct decision” to be when the RNNs DV for the correct response was greater than the other DV and the larger DV was greater than a pre-set threshold of 0.6. We evaluated the network 500ms before the checkerboard was turned off (the end of the trial). We required this criteria to be satisfied for at least 65% of both leftward and rightward trials. We note that this only affected how we terminated training. It had no effect on the backpropagated gradients, which depended on the mean-squared-error loss function. Note that a trial that outputted the correct target but did not reach the 0.6 threshold would not be counted towards the 65% criteria.

When testing, we defined the RNNs decision to be either: (1) whichever DV output (for left or right) first crossed a pre-set threshold of 0.6, or (2) if no DV output crossed the pre-set threshold of 0.6 by the end of the ‘Decision epoch,’ then the decision was for whichever DV had a higher value at the end of this epoch — an approach that is well established in models of decision-making47,48. If the RNN’s decision on a single trial was the same as the correct response, we labeled this trial ‘correct.’ Otherwise, it was incorrect. The proportion of decisions determined under criterion (2) was negligible (0.5% across 100 trials for each of 28 conditions). An interpretation for criterion (2) is that if the RNN’s DV has not achieved the threshold certainty level by the end of a trial, we assign the RNN’s decision to be the direction for which its DV had the largest value. Finally, in training only, we introduced ‘catch’ trials 10% of the time. On 50% of catch trials, no inputs were shown to the RNN and zdes(t) = 0 for all t. On the remaining 50% of catch trials, the targets were shown to the RNN, but no coherence information was shown; likewise, zdes(t) = 0 for all t on these catch trials.

We trained the three-area RNNs by constraining the recurrent weight matrix Wrec to have connections between the first and second areas and the second and third areas. In a multi-area network with N neurons and m areas, each area had N/m neurons. In our 3-area networks, each area had 100 units. Of these 100 units, 80 were excitatory and 20 were inhibitory. Excitatory units were constrained to have only positive outgoing weights, while inhibitory units were constrained to have only negative outgoing weights. We used the pycog repository25 to implement these architecture constraints. The parameters for the exemplar RNN used in the paper are shown in Table 1. In our hyperparameter sweeps, we varied the hyperparameters of the exemplar RNN. For each parameter configuration, we trained 8 different networks with different random number generator seeds.

Additional description of analyses

Decoding analysis for DLPFC and PMd data

For DLPFC and PMd data, we calculated decoding accuracy using 400 ms bins. We report numbers in a window [-300ms, +100 ms] aligned to movement onset. We used the Python sklearn.svm.SVC command with 80% training and 20 % test sets. Decoding analyses were performed using 2-49 simultaneously recorded units from Plexon U-probes and the averages reported are across PMd and DLPFC sessions, respectively. To assess whether decoding accuracies were significant, we shuffled the trials for 100 times. The decoding accuracy for direction, color and target configuration variables of a session was judged to be significant if it lies above the 99 percentile of shuffled accuracy. For DLPFC, direction, target configuration and color decoding accuracy of 100%, 78%, 53% sessions were judged to be significantly above chance; for PMd, 97%, 26%, 20% sessions demonstrated significant decoding accuracy to direction, target configuration and color.

Mutual information was calculated by computing H(Y) − LCE where H(Y) was 1 and LCE denotes the cross entropy loss (in bits). We computed the decoding and information only for sessions with decoding accuracy significantly above chance. Negative mutual information was set to zero.

Decoding and Mutual information for RNNs

We used a decoder and mutual information approximation to quantify the amount of information (color, target configuration, direction) present in the network. We trained a neural network to predict a relevant choice (for example, color) on a test set from the activity of a population of units. We used 700 trials for training, and 2100 independent trials for testing. To generate the trials for training and testing, we increased the recurrent noise to be drawn from the distribution (𝒩 (0, 0.12)) to prevent overfitting. For each trial, we averaged data in a window [-300ms, +100ms] around reaction time.

We trained a neural network with 3 layers, 64 units per layer, leakyRelu activation (α=0.2), and dropout (p=0.5), using stochastic gradient descent, to predict the choice given the activity of the population. We removed the leakyRelu activation for the linear network, and increased dropout (p=0.8). For both the nonlinear and linear network, we trained the neural network to minimize the cross-entropy loss. We used the same neural network from the decode to compute an approximation to mutual information, described in Supplementary Note 2.

RNN behavior

To evaluate the RNN’s psychometric curve and reaction-time behavior, we generated 200 trials for each of the 28 conditions, producing 400 trials for each signed coherence. For these trials, we calculated the proportion of red decisions by the RNN. This corresponds to all trials where the DV output for the red target first crossed the preset threshold of 0.6; or, if no DV output crossed the threshold of 0.6, if the DV corresponding to the red target exceeded that corresponding to the green target. The reaction time was defined to be the time between checkerboard onset to the first time a DV output exceeded the preset threshold of 0.6. If the DV output never exceeded a threshold of 0.6, in the reported results, we did not calculate a RT for this trial.

dPCA

Demixed principal components analysis (dPCA) is a dimensionality reduction technique that provides a projection of the data onto task related dimensions while preserving overall variance49. dPCA achieves these aims by minimizing a loss function: Here, Xc refers to data averaged over a “dPCA condition” (such as time, coherence, target configuration, color, or direction), having the same shape as X ∈ ℝN ×cT, but with the entries replaced with the condition-averaged response. The aim is to recover (per dPCA condition c) a Pc and Dc matrix. Pc is constrained to have orthonormal columns, while Dc is unconstrained. The number of columns of Pc and rows of Dc reflects the number of components one seeks to find per condition. The column of Pc reflects how much the demixed data contributes to each neuron. We use the principal axes from Pc to compute the axis overlap, as in Kobak et al49. We used axes of dimension 1 for RNNs, which were sufficient to capture most color, target configuration, or direction variance. For the neural data, we used five components for direction, color and target configuration since the PMd data was higher dimensional than the RNNs.

For multi-area analyses, we separated the units for each area and found the task-relevant axes for this subset of units. For the inter-area analyses, we used RNNs with only excitatory connections, and therefore found the color and direction axis using only the excitatory units. In all other analyses, all units were used to identify the axes. For RNN activity, we performed dPCA using activity over the entire trial.

For neural data, due to the stochasticity in task design, there is a trial-by-trial difference in interval between target and checkerboard onset (TC interval). The reaction time (from the checkerboard onset to monkey’s hand movement initiation) also varies for each trial. To align time events across trials, we restretched the firing rates in each trial. For DLPFC units, each trial was aligned to targets onset first. Median reaction time (527ms) and TC interval (735ms) were calculated by combining every trial in the database. For each trial, TC interval and reaction time was either compressed or stretched to the median values through linear interpolation. After the data restretching, we choose the data window T as 1300ms, from -100ms to 1200ms around target onset with sample size of 1ms. For every unit n in total units number N, we averaged the single-trial firing rate by stimulus S (checkerboard dominant color, green or red) and decision of choice D (left or right). As a result, a 4D firing-rate matrix XN×S×D×T was created as input to demixed principal component analysis algorithm.

For PMd units, activities before checkerboard onset were minimal. As a result, each trial was aligned to target onset and a segment with time window of [−100ms, 367ms] was chosen first. Then the same trial was aligned to checkerboard first and a segment with a window of [− 368ms, 465ms] was chosen. The final restretched data was the concatenation of these two data segments.

When computing the overlap in Fig. 4, we averaged across 8 initializations, and computed the PSTHs over 700 trials. In our dpca variance sweeps (Fig. 5), we computed the PSTHs over 280 trials.

PCA

Principal components analysis (PCA) is a dimensionality reduction technique that projects high-dimension data into low-dimensional axis which maximize the variance in the data. PCA provides low-dimensional projections by minimizing the loss function: XN×T is high-dimension raw data and DN×N is the decoding matrix. The low-dimension trajectories xM×T (M < N) calculated by multiplying first M rows of D by X.

Before applying PCA on the data, the raw data was preprocessed by data normalization and average firing rate removal:

Data normalization for PCA

Due to the hererogenity in firing rates, PCA results might be dominanted by units with high firing rates. To equalize the contribution of each unit in PCA, for each unit n ∈ 𝒩, 99 percentile of max firing rate xn,99 was calculated over all stimulus, decision conditions and time point. Then we divided each data for unit xn,99 by square root of to normalize the data.

Condition independent signal removal for PCA

The condition independent signal is another source that explains substantial amount of population variance other than task-related signal. Before conducting principal component analysis (PCA), we calculated the average firing rate of each single unit X1×1×1×T over all stimulus and decision conditions and subtracted this condition independent signal from the time-restretched data X1×S×D×T.

Canonical correlation

We applied CCA to assess the similarity between neural activity and the artificial unit activity50. Before applying CCA, we performed principal component analysis to reduce the dimensionality of the artificial and neural activity to remove noise, which can be arbitrarily reshaped to increase the canonical correlation50. We reduced the dimensionality of PMd data to 2 (which captures over 80% of the PMd variance). For DLPFC, 18 dimensions were required to capture over 80% of the variance, but at such a high dimensionality, noise can be reshaped to significantly increase the canonical correlations. For DLPFC, we therefore show a comparison to the top 4 PCs in Fig. 3e. However, the trends held irrespective of the DLPFC dimensionality we chose, as shown in Fig. S3. In all cases, we compared to CCA with the number of dimensions equal to 2, looking at the We report the average CCA correlation coefficients in Fig. 3e using times in a window of [-400ms, 400ms] aligned to checkerboard onset. The data was binned in 10ms bins.

Analyses of inputs and activity

In order to disentangle the effects of external inputs and recurrence, in Fig. S6, we evaluated the input contribution and overall activity. For Area 1, we defined the input contribution as Winut, and for areas 2 and 3, we defined the input contribution as , and respectively, where denotes the activity of the units in area m. The activity rm corresponds to the firing rate that experimentalists could measure, reflecting a combination of input and recurrent interactions. For constant inputs, a stable value of the activity implies there is little recurrent processing.

Inter-Area Projection Analyses

To calculate the overlap between the color and direction axes with the potent and null spaces, we performed singular value decomposition on the inter-area connections, W21 and W32. W21 and W32 were 80 × 80 matrices, and were full rank. Nevertheless, they had near some zero singular values, indicating that the effective rank of the matrix was less than 80. We defined the potent dimensions to be the top m right singular vectors, while the null dimensions were the remaining 80 − m right singular vectors.

We performed the analyses of Fig. 4f by varying the potent and null dimensions, sweeping m from 1 to 80. For each defined potent and null space, we calculated the axis overlap between the direction (or color) axis and the potent (or null) space by computing the L2-norm of the orthogonal projection (squared). We report the squared quantity because the expectation of the norm of a projection of a random vector onto an m-dimensional subspace of an n-dimensional space is m/n. We include an approximation of the expectation of the projection of a random vector in Fig. 4 by averaging the projection of 100 random vectors. Our results show that the direction axis was always more aligned with potent dimensions than the color axis, irrespective of the choice of m, and that the direction axis was preferentially aligned with the top singular vector.

eLife Assessment

This manuscript reports a useful computational study of information encoding across the monkey prefrontal and pre-motor cortices during decision making. While many of the conclusions are supported with solid analyses, the evidence for the main claim, the role of an information bottleneck across areas, is incomplete. Refocusing the paper as an RNN modeling study would increase its appeal to a systems and computational neuroscience audience.

Reviewer #1 (Public Review):

In this study, the authors aim to understand why decision formation during behavioural tasks is distributed across multiple brain areas. They hypothesize that multiple areas are used in order to implement an information bottleneck (IB). Using neural activity recorded from monkey DLPFC and PMd performing a 2-AFC task, they show that DLPFC represents various task variables (decision, color, target configuration), while downstream PMd primarily represents decision information. Since decision information is the only information needed to make a decision, the authors point out that PMd has a minimal sufficient representation (as expected from an IB). They then train 3-area RNNs on the same task and show that activity in the first and third areas resemble the neural representations of DLPFC and PMd, respectively. In order to propose a mechanism, they analyse the RNN and find that area 3 ends up with primarily decision information because feedforward connections between areas primarily propagate decision information.

The paper addresses a deep, normative question, namely why task information is distributed across several areas.

Overall, it reads well and the analysis is well done and mostly correct (see below for some comments). My major problem with the paper is that I do not see that it actually provides an answer to the question posed (why is information distributed across areas?). I find that the core problem is that the information bottleneck method, which is evoked throughout the paper, is simply a generic compression method. Being a generic compressor, the IB does not make any statements about how a particular compression should be distributed across brain areas - see major points (1) and (2).

If I ignore the reference to the information bottleneck and the question of why pieces of information are distributed, I still see a more mechanistic study that proposes a neural mechanism of how decisions are formed, in the tradition of RNN-modelling of neural activity as in Mante et al 2013. Seen through this more limited sense, the present study succeeds at pointing out a good model-data match. I point out some suggestions for improvement below.

Major points

(1) It seems to me that the author's use of the IB is based on the reasoning that deep neural networks form decisions by passing task information through a series of transformations/layers/areas and that these deep nets have been shown to implement an IB. Furthermore, these transformations are also loosely motivated by the data processing inequality.

However, assuming as a given that deep neural networks implement an IB does not mean that an IB can only be implemented through a deep neural network. In fact, IBs could be performed with a single transformation just as well. More formally, a task associates stimuli (X) with required responses (Y), and the IB principle states that X should be mapped to a representation Z, such that I(X;Z) is minimal and I(Y,Z) is maximal. Importantly, the form of the map Z=f(X) is not constrained by the IB. In other words, the IB does not impose that there needs to be a series of transformations. I therefore do not see how the IB by itself makes any statement about the distribution of information across various brain areas.

A related problem is that the authors really only evoke the IB to explain the representation in PMd: Fig 2 shows that PMd is almost only showing decision information, and thus one can call this a minimal sufficient representation of the decision (although ignoring substantial condition independent activity). However, there is no IB prediction about what the representation of DLPFC should look like. Consequently, there is no IB prediction about how information should be distributed across DLPFC and PMd.

(2) Now the authors could change their argument and state that what is really needed is an IB with the additional assumption that transformations go through a feedforward network. However, even in this case, I am not sure I understand the need for distributing information in this task. In fact, in both the data and the network model, there is a nice linear readout of the decision information in dPFC (data) or area 1 (network model). Accordingly, the decision readout could occur at this stage already, and there is absolutely no need to tag on another area (PMd, area 2+3).

Similarly, I noticed that the authors consider 2,3, and 4-area models, but they do not consider a 1-area model. It is not clear why the 1-area model is not considered. Given that e.g. Mante et al, 2013, manage to fit a 1-area model to a task of similar complexity, I would a priori assume that a 1-area RNN would do just as well in solving this task.

I think there are two more general problems with the author's approach. First, transformations or hierarchical representations are usually evoked to get information into the right format in a pure feedforward network. An RNN can be seen as an infinitely deep feedforward network, so even a single RNN has, at least in theory, and in contrast to feedforward layers, the power to do arbitrarily complex transformations. Second, the information coming into the network here (color + target) is a classical xor-task. While this task cannot be solved by a perceptron (=single neuron), it also is not that complex either, at least compared to, e.g., the task of distinguishing cats from dogs based on an incoming image in pixel format.

(3) I am convinced of the author's argument that the RNN reproduces key features of the neural data. However, there are some points where the analysis should be improved.

(a) It seems that dPCA was applied without regularization. Since dPCA can overfit the data, proper regularization is important, so that one can judge, e.g., whether the components of Fig.2g,h are significant, or whether the differences between DLPFC and PMd are significant.

(b) I would have assumed that the analyses performed on the neural data were identical to the ones performed on the RNN data. However, it looked to me like that was not the case. For instance, dPCA of the neural data is done by restretching randomly timed trials to a median trial. It seemed that this restretching was not performed on the RNN. Maybe that is just an oversight, but it should be clarified. Moreover, the decoding analyses used SVC for the neural data, but a neural-net-based approach for the RNN data. Why the differences?

(4) The RNN seems to fit the data quite nicely, so that is interesting. At the same time, the fit seems somewhat serendipitous, or at least, I did not get a good sense of what was needed to make the RNN fit the data. The authors did go to great lengths to fit various network models and turn several knobs on the fit. However, at least to me, there are a few (obvious) knobs that were not tested.

First, as already mentioned above, why not try to fit a single-area model? I would expect that a single area model could also learn the task - after all, that is what Mante et al did in their 2013 paper and the author's task does not seem any more complex than the task by Mante and colleagues.

Second, I noticed that the networks fitted are always feedforward-dominated. What happens when feedforward and feedback connections are on an equal footing? Do we still find that only the decision information propagates to the next area? Quite generally, when it comes to attenuating information that is fed into the network (e.g. color), then that is much easier done through feedforward connections (where it can be done in a single pass, through proper alignment or misalignment of the feedforward synapses) than through recurrent connections (where you need to actively cancel the incoming information). So it seems to me that the reason the attenuation occurs in the inter-area connections could simply be because the odds are a priori stacked against recurrent connections. In the real brain, of course, there is no clear evidence that feedforward connections dominate over feedback connections anatomically.

More generally, it would be useful to clarify what exactly is sufficient:

(a) the information distribution occurs in any RNN, i.e., also in one-area RNNs

(b) the information distribution occurs when there are several, sparsely connected areas

(c) the information distribution occurs when there are feedforward-dominated connections between areas

Reviewer #2 (Public Review):

Kleinman and colleagues conducted an analysis of two datasets, one recorded from DLPFC in one monkey and the other from PMD in two monkeys. They also performed similar analyses on trained RNNs with various architectures.

The study revealed four main findings. (1) All task variables (color coherence, target configuration, and choice direction) were found to be encoded in DLPFC. (2) PMD, an area downstream of PFC, only encoded choice direction. (3) These empirical findings align with the celebrated 'information bottleneck principle,' which suggests that FF networks progressively filter out task-irrelevant information. (4) Moreover, similar results were observed in RNNs with three modules.

While the analyses supporting results 1 and 2 were convincing and robust, I have some concerns and recommendations regarding findings 3 and 4, which I will elaborate on below. It is important to note that findings 2 and 4 had already been reported in a previous publication by the same authors (ref. 43).

Major recommendation/comments:

The interpretation of the empirical findings regarding the communication subspace in relation to the information bottleneck theory is very interesting and novel. However, it may be a stretch to apply this interpretation directly to PFC-PMd, as was done with early vs. late areas of a FF neural network.

In the RNN simulations, the main finding indicates that a network with three or more modules lacks information about the stimulus in the third or subsequent modules. The authors draw a direct analogy between monkey PFC and PMd and Modules 1 and 3 of the RNNs, respectively. However, considering the model's architecture, it seems more appropriate to map Area 1 to regions upstream of PFC, such as the visual cortex, since Area 1 receives visual stimuli. Moreover, both PFC and PMd are deep within the brain hierarchy, suggesting a more natural mapping to later areas. This contradicts the CCA analysis in Figure 3e. It is recommended to either remap the areas or provide further support for the current mapping choice.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public Review):

In this study, the authors aim to understand why decision formation during behavioural tasks is distributed across multiple brain areas. They hypothesize that multiple areas are used in order to implement an information bottleneck (IB). Using neural activity recorded from monkey DLPFC and PMd performing a 2-AFC task, they show that DLPFC represents various task variables (decision, color, target configuration), while downstream PMd primarily represents decision information. Since decision information is the only information needed to make a decision, the authors point out that PMd has a minimal sufficient representation (as expected from an IB). They then train 3-area RNNs on the same task and show that activity in the first and third areas resemble the neural representations of DLPFC and PMd, respectively. In order to propose a mechanism, they analyse the RNN and find that area 3 ends up with primarily decision information because feedforward connections between areas primarily propagate decision information.

The paper addresses a deep, normative question, namely why task information is distributed across several areas.

Overall, it reads well and the analysis is well done and mostly correct (see below for some comments). My major problem with the paper is that I do not see that it actually provides an answer to the question posed (why is information distributed across areas?). I find that the core problem is that the information bottleneck method, which is evoked throughout the paper, is simply a generic compression method.

Being a generic compressor, the IB does not make any statements about how a particular compression should be distributed across brain areas - see major points (1) and (2).

If I ignore the reference to the information bottleneck and the question of why pieces of information are distributed, I still see a more mechanistic study that proposes a neural mechanism of how decisions are formed, in the tradition of RNN-modelling of neural activity as in Mante et al 2013. Seen through this more limited sense, the present study succeeds at pointing out a good model-data match, and I could support a publication along those lines. I point out some suggestions for improvement below.

We thank the reviewer for their comments, feedback and suggestions. We are glad to hear you support the good model-data match for this manuscript. With your helpful comments, we have clarified the connections to the information bottleneck principle and also contrasted it against the information maximization principle (the InfoMax principle), an alternative hypothesis. We elaborate on these issues in response to your points below, particularly major points (1) and (2). We also address all your other comments below.

Major points

(1) It seems to me that the author's use of the IB is based on the reasoning that deep neural networks form decisions by passing task information through a series of transformations/layers/areas and that these deep nets have been shown to implement an IB. Furthermore, these transformations are also loosely motivated by the data processing inequality.

On Major Point 1 and these following subpoints, we first want to make a high-level statement before delving into a detailed response to your points as it relates to the information bottleneck (IB). We hope this high-level statement will provide helpful context for the rest of our point-by-point responses.

We want to be clear that we draw on the information bottleneck (IB) principle as a general principle to explain why cortical representations differ by brain area. The IB principle, as applied to cortex, is only stating that a minimal sufficient representation to perform the task is formed in cortex, not how it is formed. The alternative hypothesis to the IB is that brain areas do not form minimal sufficient representations. For example, the InfoMax principle states that each brain area stores information about all inputs (even if they’re not necessary to perform the task). InfoMax isn’t unreasonable: it’s possible that storing as much information about the inputs, even in downstream areas, can support flexible computation and InfoMax also supports redundancy in cortical areas. Indeed, many studies claim that action choice related signals are in many cortical areas, which may reflect evidence of an InfoMax principle in action for areas upstream of PMd.

While we observe an IB in deep neural networks and cortex in our perceptual decision-making task, we stress that its emergence across multiple areas is an empirical result. At the same time, multiple areas producing an IB makes intuitive sense: due to the data processing inequality, successive transformations typically decrease the information in a representation (especially when, e.g., in neural networks, every activation passes through the Relu function, which is not bijective). Multiple areas are therefore a sufficient and even ‘natural’ way to implement an IB, but multiple areas are not necessary for an IB. That we observe an IB in deep neural networks and cortex emerge through multi-area computation is empirical, and, contrasting InfoMax, we believe it is an important result of this paper.

Nevertheless, your incisive comments have helped us to update the manuscript that when we talk about the IB, we should be clear that the alternative hypothesis is non-minimal representations, a prominent example of which is the InfoMax principle. We have now significantly revised our introduction to avoid this confusion. We hope this provides helpful context for our point-by-point replies, below.

However, assuming as a given that deep neural networks implement an IB does not mean that an IB can only be implemented through a deep neural network. In fact, IBs could be performed with a single transformation just as well. More formally, a task associates stimuli (X) with required responses (Y), and the IB principle states that X should be mapped to a representation Z, such that I(X;Z) is minimal and I(Y,Z) is maximal. Importantly, the form of the map Z=f(X) is not constrained by the IB. In other words, the IB does not impose that there needs to be a series of transformations. I therefore do not see how the IB by itself makes any statement about the distribution of information across various brain areas.

We agree with you that an IB can be implemented in a single transformation. We wish to be clear that we do not intend to argue necessity: that multiple areas are the only way to form minimal sufficient representations. Rather, multiple areas are sufficient to induce minimal sufficient representations, and moreover, they are a natural and reasonably simple way to do so. By ‘natural,’ we mean that minimal sufficient representations empirically arise in systems with multiple areas (more than 2), including deep neural networks and the cortex at least for our task and simulations. For example, we did not see minimal sufficient representations in 1- or 2-area RNNs, but we did see them emerge in RNNs with 3 areas or more. One potential reason for this result is that sequential transformations through multiple areas can never increase information about the input; it can only maintain or reduce information due to the data processing inequality.

Our finding that multiple areas facilitate IBs in the brain is therefore an empirical result: like in deep neural networks, we observe the brain has minimal sufficient representations that emerge in output areas (PMd), even as an area upstream (DLPFC) is not minimal. While the IB makes a statement that this minimal sufficient representation emerges, to your point, the fact that it emerges over multiple areas is not a part of the IB – as you have pointed out, the IB doesn’t state where or how the information is discarded, only that it is discarded. Our RNN modeling later proposes one potential mechanism for how it is discarded. We updated the manuscript introduction to make these points:

“An empirical observation from Machine Learning is that deep neural networks tend to form minimal sufficient representations in the last layers. Although multi-layer computation is not necessary for an IB, they provide a sufficient and even “natural” way to form an IB. A representation z = f(x) cannot contain more information than the input x itself due to the data processing inequality[19]. Thus, adding additional layers typically results in representations that contain less information about the input.”

And later in the introduction:

“Consistent with these predictions of the IB principle, we found that DLPFC has information about the color, target configuration, and direction. In contrast, PMd had a minimal sufficient representation of the direction choice. Our recordings therefore identified a cortical IB. However, we emphasize the IB does not tell us where or how the minimal sufficient representation is formed. Instead, only our empirical results implicate DLPFC-PMd in an IB computation. Further, to propose a mechanism for how this IB is formed, we trained a multi-area RNN to perform this task. We found that the RNN faithfully reproduced DLPFC and PMd activity, enabling us to propose a mechanism for how cortex uses multiple areas to compute a minimal sufficient representation.”

In the context of our work, we want to be clear the IB makes these predictions:

Prediction 1: There exists a downstream area of cortex that has a minimal and sufficient representation to perform a task (i.e.,. I(X;Z) is minimal while preserving task information so that I(Z;Y) is approximately equal to I(X;Y)). We identify PMd as an area with a minimal sufficient representation in our perceptual-decision-making task.

Prediction 2 (corollary if Prediction 1 is true): There exists an upstream brain area that contains more input information than the minimal sufficient area. We identify DLPFC as an upstream area relative to PMd, which indeed has more input information than downstream PMd in our perceptual decision-making task.

Note: as you raise in other points, it could have been possible that the IB is implemented early on, e.g., in either the parietal cortex (dorsal stream) or inferotemporal cortex (ventral stream), so that DLPFC and PMd both contained minimal sufficient representations. The fact that it doesn’t is entirely an empirical result from our data. If DLPFC had minimal sufficient representations for the perceptual decision making task, we would have needed to record in other regions to identify brain areas that are consistent with Prediction 2. But, empirically, we found that DLPFC has more input information relative to PMd, and therefore the DLPFC-PMd connection is implicated in the IB process.

What is the alternative hypothesis to the IB? We want to emphasize: it isn’t single-area computation. It’s that the cortex does not form minimal sufficient representations. For example, an alternative hypothesis (“InfoMax”) would be for all engaged brain areas to form representations that retain all input information. One reason this could be beneficial is because each brain area could support a variety of downstream tasks. In this scenario, PMd would not be minimal, invalidating Prediction 1. However, this is not supported by our empirical observations of the representations in PMd, which has a minimal sufficient representation of the task. We updated our introduction to make this clear:

“But cortex may not necessarily implement an IB. The alternative hypothesis to IB is that the cortex does not form minimal sufficient representations. One manifestation of this alternative hypothesis is the “InfoMax” principle, where downstream representations are not minimal but rather contain maximal input information22. This means information about task inputs not required to perform the task are present in downstream output areas. Two potential benefits of an InfoMax principle are (1) to increase redundancy in cortical areas and thereby provide fault tolerance, and (2) for each area to support a wide variety of tasks and thereby improve the ability of brain areas to guide many different behaviors. In contrast to InfoMax, the IB principle makes two testable predictions about cortical representations. Prediction 1: there exists a downstream area of cortex that has a minimal and sufficient representation to perform a task (i.e., I(X; Z) is minimal while preserving task information so that I(Z; Y) ≈ I(X; Y)). Prediction 2 (corollary if Prediction 1 is true): there exists an upstream area of cortex that has more task information than the minimal sufficient area.”

Your review helped us realize we should have been clearer in explaining that these are the key predictions of the IB principle tested in our paper. We also realized we should be much clearer that these predictions aren’t trivial or expected, and there is an alternative hypothesis. We have re-written the introduction of our paper to highlight that the key prediction of the IB is minimal sufficient representations for the task, in contrast to the alternative hypothesis of InfoMax.

A related problem is that the authors really only evoke the IB to explain the representation in PMd: Fig 2 shows that PMd is almost only showing decision information, and thus one can call this a minimal sufficient representation of the decision (although ignoring substantial condition independent activity).

However, there is no IB prediction about what the representation of DLPFC should look like.

Consequently, there is no IB prediction about how information should be distributed across DLPFC and PMd.

We agree: the IB doesn’t tell us how information is distributed, only that there is a transformation that eventually makes PMd minimal. The fact that we find input information in DLPFC reflects that this computation occurs across areas, and is an empirical characterization of this IB in that DLPFC has direction, color and context information while PMd has primarily direction information. To be clear: only our empirical recordings verified that the DLPFC-PMd circuit is involved in the IB. As described above, if not, we would have recorded even further upstream to identify an inter-areal connection implicated in the IB.

We updated the text to clearly state that the IB predicts that an upstream area’s activity should contain more information about the task inputs. We now explicitly describe this in the introduction, copy and pasted again here for convenience.

“In contrast to InfoMax, the IB principle makes two testable predictions about cortical representations. Prediction 1: there exists a downstream area of cortex that has a minimal and sufficient representation to perform a task (i.e., I(X; Z) is minimal while preserving task information so that I(Z; Y) ≈ I(X; Y)). Prediction 2 (corollary if Prediction 1 is true): there exists an upstream area of cortex that has more task information than the minimal sufficient area.

Consistent with the predictions of the IB principle, we found that DLPFC has information about the color, target configuration, and direction. In contrast, PMd had a minimal sufficient representation of the direction choice. Our recordings therefore identified a cortical IB. However, we emphasize the IB does not tell us where or how the minimal sufficient representation is formed. Instead, only our empirical results implicate DLPFC-PMd in an IB computation Further, to propose a mechanism for how this IB is formed, we trained a multi-area RNN to perform this task.”

The only way we knew DLPFC was not minimal was through our experiments. Please also note that the IB principle does not describe how information could be lost between areas or layers, whereas our RNN simulations show that this may occur through preferential propagation of task-relevant information with respect to the inter-area connections.

(2) Now the authors could change their argument and state that what is really needed is an IB with the additional assumption that transformations go through a feedforward network. However, even in this case, I am not sure I understand the need for distributing information in this task. In fact, in both the data and the network model, there is a nice linear readout of the decision information in dPFC (data) or area 1 (network model). Accordingly, the decision readout could occur at this stage already, and there is absolutely no need to tag on another area (PMd, area 2+3).

Similarly, I noticed that the authors consider 2,3, and 4-area models, but they do not consider a 1-area model. It is not clear why the 1-area model is not considered. Given that e.g. Mante et al, 2013, manage to fit a 1-area model to a task of similar complexity, I would a priori assume that a 1-area RNN would do just as well in solving this task.

While decision information could indeed be read out in Area 1 in our multi-area model, we were interested in understanding how the network converged to a PMd-like representation (minimal sufficient) for solving this task. Empirically, we only observed a match between our model representations and animal cortical representations during this task when considering multiple areas. Given that we empirically observed that our downstream area had a minimal sufficient representation, our multi-area model allowed how this minimal sufficient representation emerged (through preferential propagation of task-relevant information).

We also analyzed single-area networks in our initial manuscript, though we could have highlighted these analyses more clearly to be sure they were not overlooked. We are clearer in this revision that we did consider a 1-area network (results in our Fig 5). While a single-area RNN can indeed solve this task, the single area model had all task information present in the representation, and did not match the representations in DLPFC or PMd. It would therefore not allow us to understand how the network converged to a PMd-like representation (minimal sufficient) for solving this task. We updated the schematic in Fig 5 to add in the single-area network (which may have caused the confusion).

We have added an additional paragraph commenting on this in the discussion. We also added an additional supplementary figure with the PCs of the single area RNN (Fig S15). We highlight that single area RNNs do not resemble PMd activity because they contain strong color and context information.

In the discussion:

“We also found it was possible to solve this task with single area RNNs, although they did not resemble PMd (Figure S15) since it did not form a minimal sufficient representation. Rather, for our RNN simulations, we found that the following components were sufficient to induce minimal sufficient representations: (1) RNNs with at least 3 areas, following Dale’s law (independent of the ratio of feedforward to feedback connections).”

I think there are two more general problems with the author's approach. First, transformations or hierarchical representations are usually evoked to get information into the right format in a pure feedforward network. An RNN can be seen as an infinitely deep feedforward network, so even a single RNN has, at least in theory, and in contrast to feedforward layers, the power to do arbitrarily complex transformations. Second, the information coming into the network here (color + target) is a classical xor-task. While this task cannot be solved by a perceptron (=single neuron), it also is not that complex either, at least compared to, e.g., the task of distinguishing cats from dogs based on an incoming image in pixel format.

An RNN can be viewed as an infinitely deep feedforward network in time. However, we wish to clarify two things. First, our task runs for a fixed amount of time, and therefore this RNN in practice is not infinitely deep in time. Second, if it were to perform an IB operation in time, we would expect to see color discriminability decrease as a function of time. Indeed, we considered this as a mechanism (recurrent attenuation, Figure 4a), but as we show in Supplementary Figure S9, we do not observe it to be the case that discriminability decreases through time. This is equivalent to a dynamical mechanism that removes color through successive transformations in time, which our analyses reject (Fig 4). We therefore rule out that an IB is implemented through time via an RNN’s recurrent computation (viewed as feedforward in time). Rather, as we show, the IB comes primarily through inter-areal connections between RNN areas. We clarified that our dynamical hypothesis is equivalent to rejecting the feedforward-in-time filtering hypothesis in the Results:

“We first tested the hypothesis that the RNN IB is implemented primarily by recurrent dynamics (left side of Fig. 4a). These recurrent dynamics can be equivalently interpreted as the RNN implementing a feedforward neural network in time.”

The reviewer is correct that the task is a classical XOR task and not as complex as e.g., computer vision classification. That said, our related work has looked at IBs for computer vision tasks and found them in deep feedforward networks (Kleinman et al., ICLR 2021). Even though the task is relatively straightforward, we believe it is appropriate for our conclusions because it does not have a trivial minimal sufficient representation: a minimal sufficient representation for XOR must contain only target, but not color or target configuration information. This can only be solved via a nonlinear computation. In this manner, we favor this task because it is relatively simple, and the minimal sufficient representations are interpretable, while at the same time not being so trivially simple (the minimal sufficient representations require nonlinearity to compute).

Finally, we want to note that this decision-making task is a logical and straightforward way to add complexity to classical animal decision-making tasks, where stimulus evidence and the behavioral report are frequently correlated. In tasks such as these, it may be challenging to untangle stimulus and behavioral variables, making it impossible to determine if an area like premotor cortex represents only behavior rather than stimulus. However, our task decorrelates both the stimulus and the behaviors.

(3) I am convinced of the author's argument that the RNN reproduces key features of the neural data. However, there are some points where the analysis should be improved.

(a) It seems that dPCA was applied without regularization. Since dPCA can overfit the data, proper regularization is important, so that one can judge, e.g., whether the components of Fig.2g,h are significant, or whether the differences between DLPFC and PMd are significant.

We note that the dPCA codebase optimizes the regularization hyperparameter through cross-validation and requires single-trial firing rates for all neurons, i.e., data matrices of the form (n_Neurons x Color x Choice x Time x n_Trials), which are unavailable for our data. We recognized that you are fundamentally asking whether differences are significant or not. We therefore believe it is possible to address this through a statistical test, described further below.

In order to test whether the differences of variance explained by task variables between DLPFC and PMd are significant, we performed a shuffle test. For this test, we randomly sampled 500 units from the DLPFC dataset and 500 units from the PMd dataset. We then used dPCA to measure the variance explained by target configuration, color choice, and reach direction (e.g., VarTrueDLPFC,Color, VarTruePMd,Color).

To test if this variance was significant, we performed the following shuffle test. We combined the PMd and DLPFC dataset into a pool of 1000 units and then randomly selected 500 units from this pool to create a surrogate PMd dataset and used the remaining 500 units as a surrogate DLPFC dataset. We then again performed dPCA on these surrogate datasets and estimated the variance for the various task variables (e.g., VarShuffledDLPFC,Color ,VarShuffledPMd,Color).

We repeated this process for 100 times and estimated a sampling distribution for the true difference in variance between DLPFC and PMd for various task variables (e.g., VarTrueDLPFC,Color - VarTruePMd,Color). At the same time, we estimated the distribution of the variance difference between surrogate PMd and DLPFC dataset for various task variables (e.g., VarShuffleDLPFC,Color - VarShufflePMd,Color).

We defined a p-value as the number of shuffles in which the difference in variance was higher than the median of the true difference and divided it by 100. Note, for resampling and shuffle tests with n shuffles/bootstraps, the lowest theoretical p-value is given as 2/n, even in the case that no shuffle was higher than the median of the true distribution. Thus, the differences were statistically significant (p < 0.02) for color and target configuration but not for direction (p=0.72). These results are reported in Figure S6 and show both the true sampling distribution and the shuffled sampling distributions.

(b) I would have assumed that the analyses performed on the neural data were identical to the ones performed on the RNN data. However, it looked to me like that was not the case. For instance, dPCA of the neural data is done by restretching randomly timed trials to a median trial. It seemed that this restretching was not performed on the RNN. Maybe that is just an oversight, but it should be clarified. Moreover, the decoding analyses used SVC for the neural data, but a neural-net-based approach for the RNN data. Why the differences?

Thanks for bringing up these points. We want to clarify that we did include SVM decoding for the multi-area network in the appendix (Fig. S4), and the conclusions are the same. Moreover, in previous work, we also found that training with a linear decoder led to analogous conclusions (Fig. 11 of Kleinman et al, NeurIPS 2021). As we had a larger amount of trials for the RNN than the monkey, we wanted to allow a more expressive decoder for the RNN, though this choice does not affect our conclusions. We clarified the text to reflect that we did use an SVM decoder.

“We also found analogous conclusions when using an SVM decoder (Fig. S4).”

dPCA analysis requires trials of equal length. For the RNN, this is straightforward to generate because we can set the delay lengths to be equal during inference (although the RNN was trained on various length trials and can perform various length trials). Animals must have varying delay periods, or else they will learn the timing of the task and anticipate epoch changes. Because animal trial lengths were therefore different, their trials had to be restretched. We clarified this in the Methods.

“For analyses of the RNN, we fixed the timing of trials, obviating the need to to restretch trial lengths. Note that while at inference, we generated RNN trials with equal length, the RNN was trained with varying delay periods.”

(4) The RNN seems to fit the data quite nicely, so that is interesting. At the same time, the fit seems somewhat serendipitous, or at least, I did not get a good sense of what was needed to make the RNN fit the data. The authors did go to great lengths to fit various network models and turn several knobs on the fit. However, at least to me, there are a few (obvious) knobs that were not tested.

First, as already mentioned above, why not try to fit a single-area model? I would expect that a single area model could also learn the task - after all, that is what Mante et al did in their 2013 paper and the author's task does not seem any more complex than the task by Mante and colleagues.

Thank you for bringing up this point. As mentioned in response to your prior point, we did analyze a single-area RNN (Fig. 5d). We updated the schematic to clarify that we analyzed a single area network. Moreover, we also added a supplementary figure to qualitatively visualize the PCs of the single area network (Fig. S15). While a single area network can solve the task, it does not allow us to study how representations change across areas, nor did it empirically resemble our neural recordings. Single-area networks contain significant color, context, and direction information. They therefore do not form minimal representations and do not resemble PMd activity.

Second, I noticed that the networks fitted are always feedforward-dominated. What happens when feedforward and feedback connections are on an equal footing? Do we still find that only the decision information propagates to the next area? Quite generally, when it comes to attenuating information that is fed into the network (e.g. color), then that is much easier done through feedforward connections (where it can be done in a single pass, through proper alignment or misalignment of the feedforward synapses) than through recurrent connections (where you need to actively cancel the incoming information). So it seems to me that the reason the attenuation occurs in the inter-area connections could simply be because the odds are a priori stacked against recurrent connections. In the real brain, of course, there is no clear evidence that feedforward connections dominate over feedback connections anatomically.

We want to clarify that we did pick feedforward and feedback connections based on the following macaque atlas, reference 27 in our manuscript:

Markov, N. T., Ercsey-Ravasz, M. M., Ribeiro Gomes, A. R., Lamy, C., Magrou, L., Vezoli, J., Misery, P., Falchier, A., Quilodran, R., Gariel, M. A., Sallet, J., Gamanut, R., Huissoud, C., Clavagnier, S., Giroud, P., Sappey-Marinier, D., Barone, P., Dehay, C., Toroczkai, Z., … Kennedy, H. (2014). A weighted and directed interareal connectivity matrix for macaque cerebral cortex. Cerebral Cortex , 24(1), 17–36.

We therefore believe there is evidence for more feedforward than feedback connections. Nevertheless, as stated in response to your next point below, we ran a simulation where feedback and feedforward connectivity were matched.

More generally, it would be useful to clarify what exactly is sufficient:

(a) the information distribution occurs in any RNN, i.e., also in one-area RNNs

(b) the information distribution occurs when there are several, sparsely connected areas

(c) the information distribution occurs when there are feedforward-dominated connections between areas

We better clarify what exactly is sufficient.

- We trained single-area RNNs and found that these RNNs contained color information; additionally two area RNNs also contained color information in the last area (Fig 5d).

- We indeed found that the minimal sufficient representations emerged when we had several areas, with Dale’s law constraint on the connectivity. When we had even sparser connections, without Dale’s law, there was significantly more color information, even at 1% feedforward connections; Fig 5a.

- When we matched the percentage of feedforward and feedback connections with Dale’s law constraint on the connectivity (10% feedforward and 10% feedback), we also observed minimal sufficient representations (Fig S9).

Together, we found that minimal sufficient representations emerged when we had several areas (3 or greater), with Dale’s law constraint on the connectivity, independent of the ratio of feedforward/feedback connections. We thank the reviewer for raising this point about the space of constraints leading to minimal sufficient representations in the late area. We clarified this in the Discussion.

“We also found it was possible to solve this task with single area RNNs, although they did not resemble PMd (Figure S15) since it did not form a minimal sufficient representation. Rather, for our RNN simulations, we found that the following components were sufficient to induce minimal sufficient representations: RNNs with at least 3 areas, following Dale’s law (independent of the ratio of feedforward to feedback connections).”

Thank you for your helpful and constructive comments!

Reviewer #2 (Public Review):

Kleinman and colleagues conducted an analysis of two datasets, one recorded from DLPFC in one monkey and the other from PMD in two monkeys. They also performed similar analyses on trained RNNs with various architectures.

The study revealed four main findings. (1) All task variables (color coherence, target configuration, and choice direction) were found to be encoded in DLPFC. (2) PMD, an area downstream of PFC, only encoded choice direction. (3) These empirical findings align with the celebrated 'information bottleneck principle,' which suggests that FF networks progressively filter out task-irrelevant information. (4) Moreover, similar results were observed in RNNs with three modules.

We thank the reviewer for their comments, feedback and suggestions, which we address below.

While the analyses supporting results 1 and 2 were convincing and robust, I have some concerns and recommendations regarding findings 3 and 4, which I will elaborate on below. It is important to note that findings 2 and 4 had already been reported in a previous publication by the same authors (ref. 43).

Note the NeurIPS paper only had PMd data and did not contain any DLPFC data. That manuscript made predictions about representations and dynamics upstream of PMd, and subsequent experiments reported in this manuscript validated these predictions. Importantly, this manuscript observes an information bottleneck between DLPFC and PMd.

Major recommendation/comments:

The interpretation of the empirical findings regarding the communication subspace in relation to the information bottleneck theory is very interesting and novel. However, it may be a stretch to apply this interpretation directly to PFC-PMd, as was done with early vs. late areas of a FF neural network.

In the RNN simulations, the main finding indicates that a network with three or more modules lacks information about the stimulus in the third or subsequent modules. The authors draw a direct analogy between monkey PFC and PMd and Modules 1 and 3 of the RNNs, respectively. However, considering the model's architecture, it seems more appropriate to map Area 1 to regions upstream of PFC, such as the visual cortex, since Area 1 receives visual stimuli. Moreover, both PFC and PMd are deep within the brain hierarchy, suggesting a more natural mapping to later areas. This contradicts the CCA analysis in Figure 3e. It is recommended to either remap the areas or provide further support for the current mapping choice.

We updated the Introduction to better clarify the predictions of the information bottleneck (IB) principle. In particular, the IB principle predicts that later areas should have minimal sufficient representations of task information, whereas upstream areas should have more information. In PMd, we observed a minimal sufficient representation of task information during the decision-making task. In DLPFC, we observed more task information, particularly more information about the target colors and the target configuration.

In terms of the exact map between areas, we do not believe or intend to claim the DLPFC is the first area implicated in the sensorimotor transformation during our perceptual decision-making task. Rather, DLPFC best matches Area 1 of our model. It is important to note that we abstracted our task so that the first area of our model received checkerboard coherence and target configuration as input (and hence did not need to transform task visual inputs). Indeed, in Figure 1d we hypothesize that the early visual areas should contain additional information, which we do not model directly in this work. Future work could model RNNs to take in an image or video input of the task stimulus. In this case, it would be interesting to assess if earlier areas resemble visual cortical areas. We updated the results, where we first present the RNN, to state the inputs explicitly and be clear the inputs are not images or videos of the checkerboard task.

“The RNN input was 4D representing the target configuration and checkerboard signed coherence, while the RNN output was 2D, representing decision variables for a left and right reach (see Methods).”

Another reason that we mapped Area 1 to DLPFC is because anatomical, physiological and lesion studies suggest that DLPFC receives inputs from both the dorsal and ventral stream (Romanski, et, al, 2007; Hoshi, et al, 2006; Wilson, at al, 1993). The dorsal stream originates from the occipital lobe, passes through the posterior parietal cortex, to DLPFC, which carries visuospatial information of the object. The ventral stream originates from the occipital lobe, passes through the inferior temporal cortex, ventrolateral prefrontal cortex to DLPFC, which encodes the identity of the object, including color and texture. In our RNN simulation, Area 1 receives processed inputs of the task: target configuration and the evidence for each color in the checkerboard. Target configuration contains information of the spatial location of the targets, which represents the inputs from the dorsal stream, while evidence for each color by analogy is the input from the ventral stream. Purely visual areas would not fit this dual input from both the dorsal and ventral stream. A potential alternative candidate would be the parietal cortex which is largely part of the dorsal stream and is thought to have modest color inputs (although there is some shape and color selectivity in areas such as LIP, e.g., work from Sereno et al.). On balance given the strong inputs from both the dorsal and ventral stream, we believe Area 1 maps better on to DLPFC than earlier visual areas.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) Line 35/36: Please specify the type of nuisance that the representation is robust to. I guess this refers to small changes in the inputs, not to changes in the representation itself.

Indeed it refers to input variability unrelated to the task. We clarified the text.

(2) For reference, it would be nice to have a tick for the event "Targets on" in Fig.2c.

In this plot, the PSTHs are aligned to the checkerboard onset. Because there is a variable time between target and checkerboard onset, there is a trial-by-trial difference of when the target was turned on, so there is no single place on the x-axis where we could place a “Targets on” tick. In response to this point, we generated a plot with both targets on and check on alignment, with a break in the middle, shown in Supplementary Figure S5.

(3) It would strengthen the comparison between neural data and RNN if the DPCA components of the RNN areas were shown, as they are shown in Fig.2g,h for the neural data.

We include the PSTHs plotted onto the dPCA components here for Area 1 of the exemplar network. Dashed lines indicate a left reach, while solid lines indicate a right reach, and the color corresponds to the color of the selected target. As expected, we find that the dPCA components capture the separation between components. We emphasize that the trajectory paths along the decoder axes are not particularly meaningful to interpret, except to demonstrate whether variables can be decoded or not (as in Fig 2g,h, comparing DLPFC and PMd). The decoder axes of dPCA are not constrained in any way, in contrast to the readout (encoder) axis (see Methods). This is why our manuscript focuses on analyzing the readout axes. However, if the reviewer strongly prefers these plots to be put in the manuscript, we will add them.

(4) The session-by-session decode analysis presented in Fig.2i suggests that DLPFC has mostly direction information while in Area 1 target information is on top, as suggested by Fig.3g. An additional decoding analysis on trial averaged neural data, i.e. a figure for neural data analogous to Fig.3g,h, would allow for a more straightforward and direct comparison between RNN and neural data.

We first clarify that we did not decode trial-averaged neural data for either recorded neural data or RNNs. In Fig 3g, h (for the RNN) all decoding was performed on single trial data and then averaged. We have revised the main manuscript to make this clear. Because of this, the mean accuracies we reported for DLPFC and PMd in the text are therefore computed in the same way as the mean accuracies presented in Fig 3g, h. We believe this likely addresses your concern: i.e., the mean decode accuracies presented for both neural data and the RNN were computed the same way.

If the above paragraph did not address your concern, we also wish to be clear that we presented the neural data as histograms, rather than a mean with standard error, because we found that accuracies were highly variable depending on electrode insertion location. For example, some insertions in DLPFC achieved chance-levels of decoding performance for color and target configuration. For this reason, we prefer to keep the histogram as it shows more information than reporting the mean, which we report in the main text. However, if the reviewer strongly prefers us to make a bar plot of these means, we will add them.

(5) Line 129 mentions an analysis of single trials. But in Fig.2i,j sessions are analyzed. Please clarify.

For each session, we decode from single trials and then average these decoding accuracies, leading to a per-session average decoding accuracy. Note that for each session, we record from different neurons. In the text, we also report the average over the sessions. We clarified this in the text and Methods.

(6) Fig.4c,f show how color and direction axes align with the potent subspaces. We assume that the target axis was omitted here because it highly aligns with the color axis, yet we note that this was not pointed out explicitly.

You are correct, and we revised the text to point this out explicitly.

“We quantified how the color and direction axis were aligned with these potent and null spaces of the intra-areal recurrent dynamics matrix of Area 1 ($\W^1_{rec}$). We did not include the target configuration axis for simplicity, since it highly aligns with the color axis for this network.”

(7) The caption of Fig.4c reads: "Projections onto the potent space of the intra-areal dynamics for each area." Yet, they only show area 1 in Fig.4c, and the rest in a supplement figure. Please refer properly.

Thank you for pointing this out. We updated the text to reference the supplementary figure.

(8) Line 300: "We found the direction axis was more aligned with the potent space and the color axis was more aligned with the null space." They rather show that the color axis is as aligned to the potent space as a random vector, but nothing about the alignments with the null space. Contrarily, on line 379 they write "...with the important difference that color information isn't preferentially projected to a nullspace...". Please clarify.

Thank you for pointing this out. We clarified the text to read: “We found the direction axis was more aligned with the potent space”. The text then describes that the color axis is aligned like a random vector: “In contrast, the color axis was aligned to a random vector.”

(9) Line 313: 'unconstrained' networks are mentioned. What constraints are implied there, Dale's law? Please define and clarify.

Indeed, the constraint refers to Dale’s law constraints. We clarified the text: “Further, we found that W21 in unconstrained 3 area networks (i.e., without Dale's law constraints) had significantly reduced…”

(10) Line 355 mentions a 'feedforward bottleneck'. What does this exactly mean? No E-I feedforward connections, or...? Please define and clarify.

This refers to sparser connections between areas than within an area, as well as a smaller fraction of E-I connections. We clarified the text to read:

“Together, these results suggest that a connection bottleneck in the form of neurophysiological architecture constraints (i.e., sparser connections between areas than within an area, as well as a smaller fraction of E-I connections) was the key design choice leading to RNNs with minimal color representations and consistent with the information bottleneck principle.”

(11) Fig.5c is supposedly without feedforward connections, but it looks like the plot depicts these connections (i.e. identical to Fig.5b).

In Figure 5, we are varying the E to I connectivity in panel B, and the E-E connectivity in panel C. We vary the feedback connections in Supp Fig. S12. We updated the caption accordingly.

(12) For reference, it would be nice to have the parameters of the exemplar network indicated in the panels of Fig.5.

We updated the caption to reference the parameter configuration in Table 1 of the Appendix.

(13) Line 659: incomplete sentence

Thank you for pointing this out. We removed this incomplete sentence.

(14) In the methods section "Decoding and Mutual information for RNNs" a linear neural net decoder as well as a nonlinear neural net decoder are described, yet it was unclear which one was used in the end.

We used the nonlinear network, and clarified the text accordingly. We obtained consistent conclusions using a linear network, but did not include these results in the text. (These are reported in Fig. 11 of Kleinman et al, 2021). Moreover, we also obtain consistent results by using an SVM decoder in Fig. S4 for our exemplar parameter configuration.

(15) In the discussion, the paragraph starting from line 410 introduces a new set of results along with the benefits of minimal representations. This should go to the results section.

We prefer to leave this as a discussion, since the task was potentially too simplistic to generate a clear conclusion on this matter. We believe this remains a discussion point for further investigation.

(16) Fig S5: hard to parse. Show some arrows for trajectories (a) (d) is pretty mysterious: where do I see the slow dynamics?

Slow points are denoted by crosses, which forms an approximate line attractor. We clarified this in the caption.

Reviewer #2 (Recommendations For The Authors):

Minor recommendations (not ordered by importance)

(1) Be more explicit that the recordings come from different monkeys and are not simultaneously recorded. For instance, say 'recordings from PFC or PMD'. Say early on that PMD recordings come from two monkeys and that PFC recordings come from 1 of those monkeys. Furthermore, I would highlight which datasets are novel and which are not. For instance, I believe the PFC dataset is a previously unpublished dataset and should be highlighted as such.

We added: “The PMd data was previously described in a study by Chandrasekaran and colleagues” to the main text which clarifies that the PMd data was previously recorded and has been analyzed in other studies.

(2) I personally feel that talking about 'optimal', as is done in the abstract, is a bit of a stretch for this simple task.

In using the terminology “optimal,” we are following the convention of IB literature that optimal representations are sufficient and minimal. The term “optimal” therefore is task-specific; every task will have its own optimal representation. We clarify in the text that this definition comes from Machine Learning and Information Theory, stating:

“The IB principle defines an optimal representation as a representation that is minimal and sufficient for a task or set of tasks.”

In this way, we take an information-theoretic view for describing multi-area representations. This view was satisfactory for explaining and reconciling the multi-area recordings and simulations for this task, and we think it is helpful to provide a normative perspective for explaining the differences in cortical representations by brain area. Even though the task is simple, it still allows us to study how sensory/perceptual information is represented, and well as how choice-related information is being represented.

(3) It is mentioned (and even highlighted) in the abstract that we don't know why the brain distributes computations. I agree with that statement, but I don't think this manuscript answers that question. Relatedly, the introduction mentions robustness as one reason why the brain would distribute computations, but then raises the question of whether there is 'also a computational benefit for distributing computations across multiple areas'. Isn't the latter (robustness) a clear 'computational benefit'?

We decided to keep the word “why” in the abstract, because this is a generally true statement (it is unclear why the brain distributes computation) that we wish to convey succinctly, pointing to the importance of studying this relatively grand question (which could only be fully answered by many studies over decades). We consider this the setting of our work. However, to avoid confusion that we are trying to give a full answer to this question, we are now more precise in the first paragraph of our introduction as to the particular questions we ask that will take a step towards this question. In particular, the first paragraph now asks these questions, which we answer in our study.

“For example, is all stimuli and decision-related information present in all brain areas, or do the cortical representations differ depending on their processing stage? If the representations differ, are there general principles that can explain why the cortical representations differ by brain area?”

We also removed the language on robustness, as we agree it was confusing. Thank you for these suggestions.

(4) Figure 2e and Fig. 3d, left, do not look very similar. I suggest zooming in or rotating Figure 2 to highlight the similarities. Consider generating a baseline CCA correlation using some sort of data shuffle to highlight the differences.

The main point of the trajectories is to demonstrate that both Area 1 and DLPFC represent both color and direction. We now clarify this in the manuscript. However, we do not intend for these two plots to be a rigorous comparison of similarity. Rather, we quantify similarity using CCA and our decoding analysis. We also better emphasize the relative values of the CCA, rather than the absolute values.

(5) Line 152: 'For this analysis, we restricted it to sessions with significant decode accuracy with a session considered to have a significant decodability for a variable if the true accuracy was above the 99th percentile of the shuffled accuracy for a session.' Why? Sounds fishy, especially if one is building a case on 'non-decodability'. I would either not do it or better justify it.

The reason to choose only sessions with significant decoding accuracy is that we consider those sessions to be the sessions containing information of task variables. In response to this comment, we also now generate a plot with all recording sessions in Supplementary Figure S7. We modified the manuscript accordingly.

“For this analysis, we restricted it to sessions with significant decode accuracy with a session considered to have a significant decodability for a variable if the true accuracy was above the 99th percentile of the shuffled accuracy for a session. This is because these sessions contain information about task variables. However, we also present the same analyses using all sessions in Fig. S7.”

(6) Line 232: 'The RNN therefore models many aspects of our physiological data and is therefore'. Many seems a stretch?

We changed “many” to “key.”

(7) The illustration in Fig. 4B is very hard to understand, I recommend removing it.

We are unsure what this refers to, as Figure 4B represents data of axis overlaps and is not an illustration.

(8) At some point the authors use IB instead of information bottleneck (eg line 288), I would not do it.

We now clearly write that IB is an abbreviation of Information Bottleneck the first time it is introduced.

(9) Fig. 5 caption is insufficient to understand it. Text in the main document does not help. I would move most part of this figure, or at least F, to supplementary. Instead, I would move the results in S11 and S10 to the main document.

We clarified the caption to summarize the key points. It now reads:

“Overall, neurophysiological architecture constraints in the form of multiple areas, sparser connections between areas than within an area, as well as a smaller fraction of E-I connections lead to a minimal color representation in the last area.”

(10) Line 355: 'Together, these results suggest that a connection bottleneck in the form of neurophysiological architecture constraints was the key design choice leading to RNNs with minimal color representations and consistent with the information bottleneck principle.' The authors show convincingly that increased sparsity leads to the removal of irrelevant information. There is an alternative model of the communication subspace hypothesis that uses low-rank matrices, instead of sparse, to implement said bottlenecks (https://www.biorxiv.org/content/10.1101/2022.07.21.500962v2)

We thank the reviewer for pointing us to this very nice paper. Indeed, a low-rank connectivity matrix is another mechanism to limit the amount of information that is passed to subsequent areas. In fact, the low-rank matrix forms a hard-version of our observations as we found that task-relevant information was preferentially propagated along the top singular mode of the inter-areal connectivity matrix. In our paper we observed this tendency naturally emerges through training with neurophysiological architecture constraints. In the paper, for the multi-area RNN, they hand-engineered the multi-area network, whereas our network is trained. We added this reference to our discussion.

Thank you for your helpful and constructive comments.

Version 2

Abstract

Abstract

Decision-making emerges from distributed computations across multiple brain areas, but it is unclear why the brain distributes the computation. In deep learning, artificial neural networks use multiple areas (or layers) and form optimal representations of task inputs. These optimal representations are sufficient to perform the task well, but minimal so they are invariant to other irrelevant variables. We recorded single neurons and multiunits in dorsolateral prefrontal cortex (DLPFC) and dorsal premotor cortex (PMd) in monkeys during a perceptual decision-making task. We found that while DLPFC represents task-related inputs required to compute the choice, the downstream PMd contains a minimal sufficient, or optimal, representation of the choice. To identify a mechanism for how cortex may form these optimal representations, we trained a multi-area recurrent neural network (RNN) to perform the task. Remarkably, DLPFC and PMd resembling representations emerged in the early and late areas of the multi-area RNN, respectively. The DLPFC-resembling area partially orthogonalized choice information and task inputs and this choice information was preferentially propagated to downstream areas through selective alignment with inter-area connections, while remaining task information was not. Our results suggest that cortex uses multi-area computation to form minimal sufficient representations by preferential propagation of relevant information between areas.

Introduction

The brain uses multiple areas to perform cognitive functions and tasks, including decision-making, multisensory integration, attention, motor control, and timing 1–10. However, we lack a principled understanding of how and why computations and representations differ by brain area. For example, is all stimuli and decision-related information present in all brain areas11,12, or do the cortical representations differ depending on their processing stage 13? If the representations differ, are there general principles that can explain why the cortical representations differ by brain area?

To answer these questions, we draw on the information bottleneck (IB) principle from Machine Learning and Information Theory. The IB principle defines an optimal representation as a representation that is minimal and sufficient for a task or set of tasks 14–16. To understand this principle, consider the binary classification task in Fig. 1a, where the task Y is to answer if an image, x, is a dog. To answer this question correctly, we do not need to store every pixel of the image, but rather can form a compressed representation of the input. More generally, for a representation z to be useful (we define a representation z to be a function of the input x, that is z = f (x)), it should be sufficient for performing the task, containing similar task information to the input itself (mathematically, I(Z; Y) ≈ I(X; Y), where I denotes the mutual information). Among all such representations, the IB defines an optimal representation to be one that is also maximally compressed, or contains minimal information I(Z; X) about the input. Such minimal sufficient representations are proven to be robust to nuisance input variability unrelated to the task, such as the color of the dog or the background 16. Although these representations contain less information than the input, they are often more useful and robust representations for solving the task 16–18.

Optimal representations are formed through an information bottleneck.

(a) Consider the task Y of discerning whether the image is of a dog. Images to the right have less information than the original image (I(Z; X) is smaller) but still contain approximately the same amount of information, I(Z; Y) to perform the task: “is this a dog?” (b) The information bottleneck trades off minimality, I(Z; X) as small as possible, with sufficiency, I(Z; Y) ≈ I(X; Y). (c) Checkerboard task. The monkey reaches to the target whose color matches the checkerboard dominant color. Because there are two equally likely target configurations where the color of the left and right targets are swapped, this task unmixes the color and direction choice. (d) A minimal sufficient representation of this task is to only retain the direction decision, in this case, reach left. A cortical information bottleneck should therefore only find direction choice information in premotor and motor output areas.

An empirical observation from Machine Learning is that deep neural networks tend to form minimal sufficient representations in the last layers. Although multi-layer computation is not necessary for an IB, they provide a sufficient and even “natural” way to form an IB. A representation z = f (x) cannot contain more information than the input x itself due to the data processing inequality 19. Thus, adding additional layers typically results in representations that contain less information about the input. This is illustrated in Fig. 1a, where different transformations of the original image decrease the mutual information I(Z; X) in the representation. But the task information contained in the representation, I(Z; Y), is similar, since all images can be used to correctly answer “is this a dog?” (Fig. 1b) In visual cortex, which also has a hierarchical structure of computations, early layers of processing have representations that contain low-level details (e.g., edges) and deeper layers represent higher-level concepts (e.g., object identity) 20,21, suggesting that these areas subserve different functions or tasks.

But cortex may not necessarily implement an IB. The alternative hypothesis to IB is that the cortex does not form minimal sufficient representations. One manifestation of this alternative hypothesis is the “InfoMax” principle, where downstream representations are not minimal but rather contain maximal input information 22. This means information about task inputs not required to perform the task are present in downstream output areas. Two potential benefits of an InfoMax principle are (1) to increase redundancy in cortical areas and thereby provide fault tolerance, and (2) for each area to support a wide variety of tasks and thereby improve the ability of brain areas to guide many different behaviors.

In contrast to InfoMax, the IB principle makes two testable predictions about cortical representations. Prediction 1: there exists a downstream area of cortex that has a minimal and sufficient representation to perform a task (i.e., I(X; Z) is minimal while preserving task information so that I(Z; Y) ≈ I(X; Y)). Prediction 2 (corollary if Prediction 1 is true): there exists an upstream area of cortex that has more task information than the minimal sufficient area.

We tested these hypotheses by combining electrophysiological recordings in behaving monkeys from multiple brain areas, and modeling using recurrent neural networks during a perceptual decision-making task. In particular, we recorded from the dorsoateral prefrontal cortex (DLPFC) and dorsal premotor cortex (PMd) as monkeys performed a decision-making task called the Checkerboard Task (Fig. 1c). In this task, the monkey discriminated the dominant color of a checkerboard composed of red and green squares and reached to a target matching the dominant color. Because the red and green target locations were randomly assigned to be left or right on each trial (“target configuration”), the direction decision is independent of the color decision. That is, a green color decision is equally likely to correspond to a left or right decision. The animal’s behavioral report was either a right or left reach, determined after combining the sensory evidence with the target configuration (Fig. 1d). While color is initially needed to solve the task, the minimal sufficient representation of the task to generate the correct output is a representation of only the direction decision without the color decision or the target configuration. The IB principle would therefore predict that (1) downstream areas should only contain the direction choice (in Fig. 1d, “reach left”) and (2) upstream areas should contain information about the task inputs and decision-making process, including the target configuration, perceived dominant color of the checkerboard, and direction choice.

Consistent with these predictions of the IB principle, we found that DLPFC has information about the color, target configuration, and direction. In contrast, PMd had a minimal sufficient representation of the direction choice. Our recordings therefore identified a cortical IB. However, we emphasize the IB does not tell us where or how the minimal sufficient representation is formed. Instead, only our empirical results implicate DLPFC-PMd in an IB computation. Further, to propose a mechanism for how this IB is formed, we trained a multi-area RNN to perform this task. We found that the RNN reproduced key features of DLPFC and PMd activity, enabling us to propose a mechanism for how cortex uses multiple areas to compute a minimal sufficient representation.

Results

The checkerboard task involves multiple brain areas

We trained monkeys to discriminate the dominant color of a central static checkerboard (15 × 15 grid) composed of red and green squares (Fig. 1c). The number of red and green squares was random on each trial, leading to different levels of discrimination difficulty. The signed color coherence linearly indicates which color is dominant in the checkerboard, with −1 corresponding to completely green, 0 to equal numbers of red and green squares, and +1 to completely red. If the monkey reached to the target matching the dominant checkerboard color, the trial was counted as a success and the monkey received a juice reward. Critically, the color decision (red or green) was decoupled from the direction decision (left or right) because the left and right target identities were random on every trial. Fig. 2a,b shows the monkey’s psychometric and reaction time (RT) behavior on this task. Monkeys made more errors and reacted more slowly for more ambiguous checkerboards compared to the almost completely red or completely green checkerboards 23.

DLPFC and PMd recordings during the checkerboard task.

(a) Psychometric and (b) reaction time curves of the monkey. X-axes in both (a) and (b) depict signed coherence which indicates the relative amount of red vs. green in the checkerboard. (c) Example DLPFC and (d) PMd PSTHs aligned to checkerboard onset. Red and green traces correspond to red and green color choices, respectively. Dotted and solid traces correspond to right and left direction choices, respectively. Data are smoothed with a 25 ms Gaussian and averaged across trials. (e) PCs 1, 3, and 4 for DLPFC and (f) PCs 1,2,3 for PMd. (g) Results of dPCA analysis for DLPFC and (h) PMd showing the dPCs for direction, target configuration, and color. (i) Histogram (across sessions) of direction, color, and target configuration decode accuracy and (j) usable information for DLPFC and PMd. The large variance in recordings is due to across-session variance.

We used linear multi-contact electrodes (U and V-probes) to record from the DLPFC (2819 single neurons and multiunits) and PMd (996 single neurons and multiunits) as the monkeys performed the checkerboard task. The PMd data was previously described in previous studies by Chandrasekaran and colleagues23,24. Example peri-stimulus time histograms (PSTHs) for neurons in DLPFC and PMd are shown in Fig. 2c, d and (and Fig. S5), respectively, where solid (dotted) lines correspond to left (right) reaches and color (red or green) denotes the color decision. DLPFC PSTHs in Fig. 2c separate based on direction choice, target configuration, and color choice, whereas PMd PSTHs primarily separate based on the direction choice, and only very modestly with target configuration or color.

Together, these examples demonstrate that DLPFC and PMd single units exhibit activity reflecting the decision-making process, implicating multiple brain areas in decision-making. Further, DLPFC likely contains multiple task-relevant signals, whereas PMd generally contains only direction choice related signals necessary for the behavioral report in the task. In the next sections, we use dimensionality reduction, decoding, and information theory to quantify the extent of color, target configuration, and direction representations in DLPFC and PMd at the population level and show that these physiological observations are consistent with the IB principle. We then use recurrent neural network models to build a mechanistic hypothesis for how an IB could be implemented.

Evidence for a cortical information bottleneck between DLPFC and PMd

Our single neuron examples suggest that neuronal responses in DLPFC are modulated by color choice and target configuration, but PMd neurons generally are not. Our hypothesis is that these cortical representations in DLPFC and PMd are consistent with the IB principle. The direct prediction of this hypothesis is that the PMd population activity should contain a minimal and sufficient representation of the behaviorally relevant output – the direction choice – while upstream DLPFC population activity should represent multiple task-relevant variables.

To study this at the population level, we performed principal components analysis (PCA) on DLPFC and PMd neural population activity. In these PCA trajectories, we subtracted the condition-independent component of the signal to better highlight representations of direction, target configuration, and color. DLPFC and PMd exhibited qualitatively different neural population trajectories (Fig. 2e,f). First, after target onset, DLPFC trajectories separated as a function of the two target configurations, and at the time of checkerboard onset (purple dots), DLPFC activity further separated into four distinct trajectories based on the four possible color × direction outcomes (green left, green right, red left, red right, Fig. 2e). Thus, DLPFC contains information about target configuration, color choice, and direction choice. Note, we chose PC 1, 2 and 4 of DLPFC as they provided a better visualization of target configuration signal between target and checkerboard. In contrast, PMd trajectories in Fig. 2f did not exhibit target-configuration-specific steady state responses. Thus, at the point of checkerboard onset (purple dots), trajectories overlapped in the top 3 principal components, and only separated based on the direction, but not color.

To quantify these these differences, we performed demixed principal component analysis (dPCA) on the DLPFC and PMd population activity (Fig. 2g,h). DLPFC and PMd activity both exhibited strong condition independent activity (82% and 86% variance, respectively). DLPFC activity represented the target configuration, but PMd did not (Fig. 2g,h, target configuration dPC). dPCA also identified principal axes that maximized variance related to the direction choice, color choice, and target configuration. In DLPFC, the top direction choice, color choice, and target configuration axes captured 7.1%, 0.9%, and 1.5% of the population activity. In PMd, these values were 10.6%, 0.2%, and 0.2%. Across all direction, color, and target configuration axes, the dPCA variance captured for DLPFC was 11%, 3%, 5%, while for PMd it was 12%, 1%, 1%. This dPCA analysis provides further evidence that DLPFC represents direction, target configuration, and color while PMd has nearly minimal representations of color and target configuration, consistent with the IB principle. The difference in color and target configuration variance explained between DLPFC and PMd is statistically significant (p < 0.02, shuffle test shown in Fig. S6).

Our dPCA results with trial-averaged firing rates suggest that axes associated with color and target configuration in PMd have very little variance associated with them. However, these results do not rule out the possibility that there is decodable information about these task-related variables on single trials. We performed two other analyses to assess if there is an IB and that PMd contains a minimal sufficient representation. We calculated the decode accuracy and an estimate of mutual information for direction, color, and target configuration in DLPFC and PMd population activity. We decoded direction, color, and target configuration from DLPFC and PMd sessions where we recorded a small population of neurons using a support vector machine (see Methods). To estimate mutual information, we quantified the Usable Information, a variational lower bound to mutual information that can be computed on high-dimensional data through estimating cross-entropy loss 17,18 (see Methods).

We found that for many DLPFC sessions we could reliably decode direction, color, and target configuration on individual trials well above chance (histogram in Fig. 2i, mean decoding accuracy across sessions: direction 86%, target configuration 59%, color 57%, for details on decoding, see Methods). To conservatively assess significance, we shuffled the trials for each session 100 times and obtained a surrogate decoding accuracy for each session, and then tested if the true accuracy was greater than the 99th percentile of the shuffled accuracy distribution. We found that 100%, 78%, 54% sessions having true accuracy higher than 99th percentile of the shuffled accuracy for direction, target configuration and color. In DLPFC, we could therefore significantly decode direction choice, target configuration, and color choice. In contrast in PMd, we rarely found sessions where target configuration and color could be reliably decoded well above chance (mean accuracy across sessions: direction 88%, target configuration 52%, color 51%), as shown in Fig. 2i. Again, we compared the true decoding accuracy with the surrogate decoding accuracy obtained by shuffling trials. We found 98%, 26%, 19% sessions having true accuracy above 99th percentile of the shuffled accuracy for direction, target configuration, and color. The differences in mean decoding accuracy in DLPFC and PMd were significant for only color and target configuration (p < 0.001, Wilcoxon rank-sum test), but not direction (p = 0.145, Wilcoxon rank-sum test). Thus, both DLPFC and PMd contain significantly more direction information but DLPFC contains more color choice and target configuration information consistent with the IB principle.

We also quantified usable information for DLPFC and PMd (Fig. 2j). For this analysis, we restricted it to sessions with significant decode accuracy with a session considered to have a significant decodability for a variable if the true accuracy was above the 99th percentile of the shuffled accuracy for a session. This is because these sessions contain information about task variables. However, we also present the same analyses using all sessions in Fig. S7. DLPFC had sessions with non-zero usable information for direction, color, and target configuration (average direction information: 0.56 bits, target configuration: 0.07 bits, color: 0.06 bits) while PMd only had non-zero usable information for direction (average direction information: 0.64 bits, target configuration: 0.013 bits, color: 0.005-bits). The differences in usable information in DLPFC and PMd were also significant for only color and target configuration (p < 0.001, Wilcoxon rank-sum test), but not direction (p = 0.107, Wilcoxon rank-sum test). Together, these results indicate that PMd had a more minimal representation of task inputs, particularly color and target configuration, than DLPFC.

Our dPCA and decoding results are consistent with the existence of a cortical IB between DLPFC and PMd that reduces the amount of target configuration and color information in PMd while preserving the direction choice information necessary to solve the task. We next modeled this multi-area IB to develop a mechanistic hypothesis for how this cortical IB could be computationally implemented.

A multi-area recurrent neural network model of DLPFC and PMd

To develop a mechanistic hypothesis for this cortical IB, we studied our previously reported multi-area RNN to perform the Checkerboard task (Fig. 3a)25. We chose to use this multi-area RNN because prior work demonstrated this RNN, like our PMd data, has a minimal color representation in Area 3. The RNN input was 4D representing the target configuration and checkerboard signed coherence, while the RNN output was 2D, representing decision variables for a left and right reach (see Methods). The RNN had 3 areas, obeyed Dale’s law 26, and had approximately 10% feedforward and 5% feedback connections between areas based on projections between prefrontal and premotor cortex in a macaque atlas27. RNN psychometric and RT curves for the multi-area RNN exhibited similar behavior to monkeys performing this task (Fig. 3b,c; across several RNNs, see Fig. S1).

RNN modeling of the CB task.

(a) Multi-area RNN configuration. The RNN received 4 inputs. The first two inputs indicated the identity of the left and right targets, which was red or green. These inputs were noiseless. The last two inputs indicated the value of the signed color coherence (proportional to amount of red in checkerboard) and negative signed color coherence (proportional to amount of green in checkerboard). We added independent Gaussian noise to these signals (see Methods). The network outputted two analog decision variables indicating evidence towards the right target (solid line) or left target (dashed line). A decision was made in the direction of whichever decision variable passed a preset threshold (0.6) first. The time at which the decision variable passed the threshold was defined to be the reaction time. (b,c) Psychometric and reaction time curves for exemplar multi-area RNN. (d) Area 1 and Area 3 principal components for exemplar RNN. (e) CCA correlation between each area and DLPFC principal components (left) and PMd principal components (right). DLPFC activity most strongly resembles Area 1, while PMd activity most strongly resembles Area 3. See also Fig. S3 where we computed CCA as a function of the number of dimensions. (f) Relative dPCA variance captured by the direction, color, and target configuration axes. Normalization makes direction variance equal to 1. Area 1 (3) variances more closely resemble DLPFC (PMd). (g) Area 1 has significantly higher decoding accuracies and (h) usable information compared to Area 3, consistent with DLPFC and PMd.

Remarkably, even though the multi-area RNN was in no way regularized to reproduce DLPFC and PMd activity, activity in Area 1 qualitatively resembled neural responses in DLPFC, representing both direction and color, while Area 3 resembled PMd, representing direction (Fig. 3d,e). Like DLPFC, Area 1 had four distinct trajectories corresponding to the four possible task outcomes and represented target configuration, direction choice, and color choice (Fig. 3d and see Fig. S2). In contrast, Area 3 population trajectories primarily separated based on direction and not by target configuration or color — remarkably similar to the trajectories observed in PMd.

We performed CCA to assess the similarity between the empirical neural trajectories to each RNN area’s neural trajectories (see Methods). The CCA analysis suggested that Area 1 exhibited the strongest resemblance to DLPFC, while Area 3 most strongly resembled PMd activity (Fig. 3e). These results show that a multi-area RNN reproduced similar behavior to the monkey, and further that it did so with architecturally and qualitatively distinct areas that resembled the physically distinct DLPFC and PMd cortical areas.

The RNN activity differs from cortical activity in two important ways. First, RNNs generally had a significantly smaller variance condition-independent signal (46.7% and 49.4% average variance in Area 1, and 3, respectively) than in DLPFC and PMd (82% and 86% variance, respectively). One possible explanation is that condition-independent variance in PMd is associated with a trigger signal, likely from the thalamus28, and these RNNs do not output arm kinematics, forces, or electromyography (EMG). Similarly, in DLPFC, we did not explicitly model the target and checkerboard inputs to have large onset signals that are often associated with visual stimulation. This significant condition independent variance in the neurophysiological data may therefore make decoding more difficult since there is relatively lower variance representing direction, color, or target configuration. While our CCA analysis was performed with the condition-independent signal removed, this difference impacts both dPCA and decoding results. In general, we found that RNN exhibited trends observed in the neurophysiological data more strongly, including more variance captured for direction, color, and target configuration, as well as higher decoding accuracies. We therefore compared the relative, rather than absolute, trends in RNN activity and DLPFC for dPCA and decoding analyses for the purposes of identifying a RNN IB similar to the cortical IB.

We found that the 3-area RNN exhibited similar trends to DLPFC and PMd activity in dPCA variance and decoding accuracy. When comparing only the top axis for direction, color, and target configuration, DLPFC activity had relatively large variance captured along the direction axis (7.1% variance captured), followed by relatively weaker representations for target configuration (1.5%) and color (0.9%). Area 1 activity had similar relative trends, with the direction axis explaining 30.9% variance, followed by target configuration (13.3%) and color (5.6%). In Fig. 3f, we show the similarity in relative trends in DLPFC and Area 1 by normalizing these quantities by the variance captured by the direction axis. Meanwhile, PMd activity exhibited more direction-related variance than DLPFC (10.6% variance captured) and did not exhibit a significant target configuration and color axis representation (0.2% for both axes). Likewise, Area 3 had a stronger representation of direction (48.5% variance captured) than in Area 1, but negligible target configuration and color axes variance (0.1% for both axes), demonstrating the same relative trend (Fig. 3f). These results show that Area 1 more strongly resembles DLPFC and Area 3 more strongly resembles PMd in relative dPCA variance.

We next evaluated the decode accuracy and usable information in the multi-area RNN using a nonlinear decoder on single trials, in line with prior work29 (see Methods). We found Area 1, like DLPFC, had significant information for direction, target configuration, and color. Direction, color, and target configuration could be decoded from Area 1 population activity at accuracies of 94.4%, 93.4%, and 99.0%, respectively, corresponding to 0.81, 0.79, and 0.92 bits of usable information (Fig. 3g,h). In contrast, Area 3 direction decode accuracy was 99.4%, while color and target configuration accuracy were significantly lower (51.1% and 54.3%, respectively). This corresponded to 0.97, 0.0023, and 0.0078 bits of usable information for direction, target configuration, and color, respectively. We also found analogous conclusions when using an SVM decoder (Fig. S4).

Together, these results show that our multi-area RNN exhibited distinct areas that resembled DLPFC and PMd activity, and also implemented an IB so that its output area only had primarily direction information and less color and target configuration information. This multi-area RNN therefore implements a candidate mechanism for how a cortical IB could be implemented between DLPFC and PMd.

Mechanistic features of the DLPFC and PMd bottleneck: partial orthogonalization and selective propagation

The multi-area RNN contains representations consistent with the IB principle — its input and output areas resemble DLPFC and PMd. The RNN therefore models key aspects of our physiological data and is therefore a candidate system to understand how multiple areas could lead to the empirically observed minimal sufficient representations. The unique advantage of the RNN is that we know the firing rates in each area as well as the within and inter-areal connections, which allows us to investigate how the multi-area RNN deemphasizes color information through an IB. We reasoned that such an IB could be implemented in three ways: Color information may be (1) primarily attenuated through recurrent neural dynamics, (2) primarily attenuated through inter-areal connections, or (3) attenuated through a combination of recurrent dynamics and inter-areal connections. This is illustrated in Fig. 4a.

IB hypotheses and mechanism.

(a) Candidate mechanisms for IB. (b) Axes overlap of the direction, color, and target configuration axes for DLPFC and RNN data. The direction axis is more orthogonal to the color and target configuration axes. (c) Projections onto the potent space of the intra-areal dynamics for Area 1 (for other areas, see Supp. Fig. S11). We computed the potent projection of the direction axis, color axis, and a random vector with each area’s intra-areal dynamics matrix. We found intra-areal dynamics amplify color information in Area 1, and do not selectively attenuate color information in Areas 2 and 3. (d) Illustration depicting how the orientation of the axes affect information propagation. Information on the direction axis (orange) can be selectively propagated through inter-areal connections which information on the color axis (maroon) is not. (e) Inter-areal hypotheses. (f) Projections onto the potent space between areas for the color axis, direction axis, and random vector. Regardless of the dimension of the potent space, the direction axis is preferentially aligned with the potent space, indicating the information along this axis propagates, while the color axis is approximately randomly aligned. We emphasize the high alignment of the direction axis: the direction axis has a stronger alignment onto the first potent dimension of W21 than the remaining dimensions combined. Meanwhile, the color axis is aligned at nearly chance levels, and will therefore be propagated significantly less than the direction axis. Shading indicates s.e.m.

To test these hypotheses, we first quantified how color, target configuration, and direction information was represented in the animals and our network. We performed dPCA in the different areas to identify demixed principal components that represented the corresponding information. In DLPFC, we quantified the overlap of the dPCA principal axes for target configuration, color, and direction. While the target configuration and color axes were relatively aligned (dot product, DP: 0.52), the direction axis was closer to orthogonal to the color axis (DP: 0.18) and the target configuration axis (DP: 0.35), as shown in Fig. 4b. These results suggest that DLPFC partially orthogonalizes information about the direction choice from the color choice and target configuration. We also observed these trends in the RNN, albeit more strongly. In DLPFC-resembling Area 1, we observed the target configuration and color were also highly aligned (DP: 0.95) but that the direction axis was more orthogonal to the color axis (DP: 0.13) and the target configuration axis (DP: 0.12). In our simulations, the reported values reflect the mean across 8 networks trained with the same hyperparameters. A candidate mechanism for this orthogonalization, found by performing dynamical analyses on the RNN, is shown in Fig. S8.

The advantage of our model is that both the intra-areal dynamics and inter-areal connectivity matrices are known. We analyzed how these axes were aligned with the intra-areal recurrent dynamics and inter-areal connectivity matrices to identify which hypothesis explained how the RNN implemented the IB. To do so, we performed singular value decomposition (SVD) on these matrices. We defined a k-dimensional “potent space” to be the right singular vectors corresponding to the k largest singular values of the matrix. The “null space” is the orthogonal complement of the potent space, which comprises the remaining d − k smallest singular vectors, where d refers to number of columns of the matrix. The null projection magnitudes are equal to one minus the potent projection. We quantified how the color and direction axis were aligned with these potent and null spaces (see Methods). This enabled us to study if the emergence of minimal sufficient representations was due to: (1) relative amplification of the direction information with respect to a random vector, (2) relative suppression of the color/target configuration information with respect to a random vector, or (3) a combination of both. Finally, we focused our analyses on Area 1 recurrent dynamics and the inter-areal connections between Areas 1 and 2 (W21) because color information is significantly attenuated by Area 2 (dPCA color variance in Area 2: 0.14%). The same analyses applied to downstream areas are shown in Fig. S11.

We first tested the hypothesis that the RNN IB is implemented primarily by recurrent dynamics (left side of Fig. 4a). These recurrent dynamics can be equivalently interpreted as the RNN implementing a feedforward neural network in time. We quantified how the color and direction axis were aligned with these potent and null spaces of the intra-areal recurrent dynamics matrix of Area 1 . We did not include the target configuration axis for simplicity, since it highly aligns with the color axis for this network. The axes found through dPCA are proxies for how task information is represented in the neural activity over the trial, while the dynamical activity during a trial depends on the complex interaction between left and right singular vectors of the recurrent matrix (and task inputs) 30. We use the alignment of the (fixed) dPCA axes with the right singular vector as a proxy for the amount of information propagated through recurrence. In Area 1, we found significant alignment of the color axis with the top right singular vectors (potent space) of the recurrent dynamics matrix (Fig. 4c). Additionally, we also performed an alternative analysis where we compared input and activity representations of color discriminability and direction discriminability for our exemplar network. We observe an amplification, not a reduction, in color discriminability with respect to the inputs in Area 1 (Fig. S9) consistent with the amplification observed in Fig. 4c. These findings argue against the hypothesis that recurrent dynamics preferentially attenuate color information by projecting it into a nullspace of the recurrent dynamics. Rather, these data suggest that Area 1 has significant color information in its potent space, indicating that the recurrent computation amplifies color information. In Areas 2 and 3, the color axis (which had small variance of 0.14% and 0.07% in Areas 2 and 3, respectively) was again typically more strongly aligned with than a random vector, (Fig. S11). In summary, the dPCA and discriminability analyses suggest that the network did not use recurrent dynamics to attenuate color information, and is therefore inconsistent with the hypothesis that the IB is primarily implemented through intra-areal recurrent dynamics.

Our alternative hypothesis is that color information is primarily attenuated through inter-areal connections. This is schematized in Fig. 4d, where inter-areal connections propagate activity along the Area 1 direction axis (orange) to Area 2, but attenuate Area 1 color axis activity (maroon). To test this hypothesis, we quantified how the color and direction axis were aligned with these potent and null spaces of the inter-areal matrices. This enabled us to quantify the alignment of the direction and color axes with the inter-areal potent and null spaces and specifically determine how direction and color information were differentially propagated (Fig. 4e). Inter-areal connections could attenuate color information by aligning the color axis with the null space of W21 (Hypothesis 1 in Fig. 4e), propagate information preferentially (Hypothesis 2 in Fig. 4e), or both attenuate and propagate information (Hypothesis 3 in Fig. 4e).

We calculated the projections for both the color and choice axes on to the potent space for the connection matrix from area 1 to area 2 (W21) The projections onto the potent space are shown in Fig. 4f for the color and direction axis. We found the direction axis was more aligned with the potent space. In fact, the direction axis was consistently most aligned with the top singular vector of the W21 matrix, on average more than the remaining d − 1 singular vectors. In contrast, the color axis was aligned to a random vector. This suggests that learning in the multi-area recurrent network involved aligning the relevant information (in the activations) with the top singular vector (governed by the learned parameters of the feedforward matrix). These results indicate that direction information is preferentially propagated to subsequent areas, while color information is aligned with a random vector. This is most consistent with the “propagate only” hypothesis shown in Fig. 4e.

Such alignment of the direction axis with the top singular vector of the connection matrix isn’t trivial: the potent space depends on the parameter W21 learned during training, while the direction axis is not a parameter but a dPCA axis computed from Area 1 activity. This alignment was robust to the dimension of the effective potent space, and was consistent across networks with varying feedforward connectivity percentages (10%, 20%, 30%, 50%, 100%). Further, we found that W21 in unconstrained 3 area networks (i.e., without Dale’s law constraints) had significantly reduced alignment of the direction axis with the top singular vectors (Fig. S11d).

In summary, the multi-area RNN IB is primarily implemented through preferential propagation of direction information through inter-areal connections. Recurrent dynamics play a role in processing color information and target configuration to arrive at direction choice information. In Area 1, RNN dynamics amplify color information. Our results are therefore most consistent with the hypothesis that the IB is implemented primarily through inter-areal connections, not recurrent dynamics, in Fig. 4a.

Effect of network architecture and training hyperparameters on the information bottleneck

We next assessed the network architectures and hyperparameters that influenced the formation of minimal sufficient representations during the Checkerboard task. We swept RNN architectural parameters and machine learning hyperparameters to assess what variables were important for learning minimal sufficient representations without color information. Specifically, we varied the connectivity type (unconstrained connectivity vs Dale’s law and varying proportion of connections), the percentage of unconstrained feedforward connections, the percentage of feedforward E to I connections, the percentage of E to E connections, the number of areas (from 1 to 4), the number of artificial networks, L2 weight regularization, L2 rate regularization, and the learning rate. Our choice of feedforward and feedback connections were informed by a macaque atlas27. In our sweeps we quantified the color (and direction) variance and accuracy in the last area.

We generally observed minimal sufficient representations in the last area so long as there was a sufficient connection bottleneck between RNN areas. In unconstrained networks, shown in Fig. 5a, color variance and decode accuracy decreased as the percentage of feedforward connections between areas decreased, though the representations were not minimal. We incorporated Dale’s law with 80% E and 20% I neurons following Song et al. 26 into subsequent sweeps (Fig. 5b-e). Minimal representations with chance color decode accuracy emerged when the percentage of feedforward E to I connections was 2 − 5% or less (the overall percentage of feedforward E to E was fixed at 10% following a macaque atlas). We also found that when there was no feedforward inhibition, but when we varied the percentage of feedforward or feedback E-to-E connections RNNs generally had nearly minimal representations (Fig. 5c, and Supp. Fig. S12). We observed that as long as there were 3 or 4 areas, there was a large decrease in color information in the last area (Fig. 5d) quantified by decoding, though note that there was a large drop in color variance for 2 area networks. These results suggest that multi-area networks, with a feedforward connection bottleneck tend to produce more minimal representations for the Checkerboard task.

Robustness of the information bottleneck across hyperparameters and computational advantage.

Varying (a) proportion of feedforward connections in an unconstrained network, (b) E-I connections in a Dale’s law network, (c) the proportion of feedforward E-E connections in a Dale’s law network; feedback connections are varied in Fig. S12, (d) the number of areas, and (e) the machine learning hyperparameters revealed that Area 3 color variance and color accuracy decrease as long as there is a connectivity bottleneck between areas. (f) Summary of these results quantifying usable information. Overall, neurophysiological architecture constraints in the form of multiple areas, sparser connections between areas than within an area, as well as a smaller fraction of E-I connections lead to a minimal color representation in the last area. Parameters of the exemplar network are in Table 1.

We also varied machine learning hyperparameters (Fig. 5e) to assess the extent to which the IB was present. To prevent an exponential search space, we fixed the architecture to the exemplar network used in this study and tested one hyperparameter at a time. We varied the number of artificial units in the network, the L2 weight regularization, the L2 rate regularization, and the learning rate. At each hyperparameter setting, we trained a total of 8 multi-area RNNs. Our exemplar network consistently exhibited little to no Area 3 color information across every hyperparameter setting we chose, suggesting that the presence of the IB is not a result of a particular choice of machine learning hyperparameters.

We again summarized all sweeps by calculating the “Usable Information”17 to quantify the direction and color information in RNNs, as shown in Fig. 5f and the results reaffirmed conclusions from the variance and decoding analyses. Together, these results suggest that a connection bottleneck in the form of neurophysiological architecture constraints (i.e., multiple areas, sparser connections between areas than within an area, as well as a smaller fraction of E-I connections) was the key design choice leading to RNNs with minimal color representations and consistent with the IB principle.

Discussion

The goal of this study was to investigate if predictions from the IB principle in machine learning and information theory are also observed in cortical circuits. The IB principle defines an optimal representation to be one that retains only the relevant or useful information for solving a task14. This principle has been applied to explain the success of deep networks 16,31, by forming minimal sufficient representations of task inputs, leading to better generalization bounds and invariance to noise 16. We explored whether such a principle could explain cortical representations across different areas during a visual perceptual decision making task by probing the ability to decode task-relevant behavioral choices and external input information from different brain areas. We found that later areas of cortex along a sensorimotor transformation (in PMd) only represented the behavioral report, that is the action choice, while earlier areas had stronger input representations and performed relevant computations to define the behavioral report (in DLPFC). To better understand how such a phenomenon could be implemented in cortex, we trained many artificial multi-area RNNs to perform this task. Surprisingly, we also observed that RNNs formed minimal sufficient representations across a range of hyperparameter settings, suggesting the formation of minimal sufficient representations may be a more general feature of multi-area computation.

Given the “full-observability” of our multi-area models, we were able to analyze the learned weight matrices and understand how the network converged to transform task inputs into minimal sufficient representations by the output area. In particular, we found that the output-relevant direction information was preferentially propagated between areas by having the largest overlap with the top singular vector of the learned feedforward matrices. In contrast, color information was almost randomly propagated through feedforward connections. This mechanism is related to prior work on output potent and output null subspaces32 and communication subspaces 1,33, with the important difference that color information isn’t preferentially projected to a nullspace, but is aligned similarly to any random vector. Preferential alignment with a cortical nullspace is therefore not necessary to achieve an IB — color information may be attenuated through random alignment to a communication subspace. This solution (random alignment) poses less constraints on inter-areal connectivity than a solution that preferentially propagates direction information while also preferentially projecting color information to a nullspace.

Our results are also consistent with recent work proposing that cortical areas convey information through communication subspaces. One observation in communication subspaces is that they do not merely propagate the directions of highest variance1. We also observed this phenomenon for the W21 connectivity matrix, which communicates information from Area 1 to Area 2. Color activity had significant variance in Area 1 (see Fig. S8). Inter-areal connections must therefore not merely propagate the highest variance dimensions of a preceding area, otherwise color information would be conveyed to Area 2. Consistent with this, we found that while the top 2 PCs capture 97.7% excitatory unit variance, the top 2 readout dimensions of W21 only captured 40.0% of Area 1’s excitatory unit neural variance (Fig. S10). Hence, inter-areal connections are not aligned with the most variable dimensions, but are rather aligned to preferentially propagate certain types of information — a result consistent with a recent study analyzing links between activity in V1 and V21.

We find minimal sufficient representations in PMd and in the later areas of our recurrent network models. Do such representations have any advantages? One possibility is that in cortex a minimal sufficient representation provides energetic benefits 34,35. Another possibility is that such a representation provides a computational advantage. This is an open question that is still somewhat unresolved in the machine learning community, with representation learning approaches that maximize mutual information between representations and inputs also leading to useful task representations36, in addition to compressed representations16,17. The information contained in the representation of a neural network is related to the “Information in the Weights” 37, which can be quantified using the Fisher Information 38–40, a measure of sensitivity to perturbations. This “Information in the Weights” view would predict that minimal sufficient representations have smaller Fisher information and are therefore less sensitive to (local) perturbations in the readout weights. In the context of deep networks, it has been proposed that minimal sufficient representations simplify the role of the output readout or classifier16. Further, a minimal sufficient representation with respect to a family of probabilistic decoders/classifiers will provably generalize better41.

Although finding a resolution to this debate in machine learning is beyond the scope of this paper, we assessed if minimal RNNs exhibited any qualities consistent with machine learning predictions. We explored whether minimal sufficient representations would simplify the readout, which we quantified by measuring the model’s performance in response to perturbations to the readout weights. We found that 3-area networks with minimal color information (particularly networks in Fig. 5b with no feedforward E-to-I connectivity) were less sensitive to perturbations than corresponding networks with significant color information (networks in Fig. 5a with 10% unconstrained feedforward connectivity, see Fig. S14). We also found that these networks differed significantly in readout complexity, with 3-area networks with minimal color information exhibiting simpler and sparser readouts (Fig. S14). However, we did not observe a clear trend between perturbation sensitivity and usable color information across random initializations (Fig. S14) for a fixed parameter setting (networks with 10% feedforward inhibition in Fig. 5b). An interesting venue for future work is to further examine the potential advantages of a minimal sufficient representation. Such findings would be valuable to the machine learning and neuroscience community. In our study, several factors including recurrent connectivity, multiple areas, and E/I populations make theoretical study of this question difficult. It is likely that studying this question requires simplifying the setting. For example, it likely makes sense to first focus on feedforward networks with a variable amount of task input information, similar to the generalized checkerboard-task used in a related study 17.

Our task could be solved with or without feedback connections with equivalent performance, indicating that feedback was not necessary to solve the task (Fig. S12). Minimal sufficient representations were found in both purely feedforward RNNs or RNNs with feedback (Fig. S12). When the model had feedback connections, we observed that feedback connections between Areas 2 and 1 preferentially conveyed direction information. Due to the presence of choice related signals in several cortical areas, these feedback connections may also play a role in computation of the direction choice. Another perspective on feedback signals is that they may related to error signals used for learning42. Multi-area networks may help understand and develop new hypotheses for physiological studies of feedforward and feedback computation 43,44, and more generally distributed processing for decision-making and cognition 8,45.Future research may use carefully designed tasks in conjunction with multi-area RNNs to better understand the role of feedback in computation.

We also found it was possible to solve this task with single area RNNs, although they did not resemble PMd (Figure S15) since it did not form a minimal sufficient representation. Rather, for our RNN simulations, we found that the following components were sufficient to induce minimal sufficient representations: RNNs with at least 3 areas, following Dale’s law (independent of the ratio of feedforward to feedback connections).

The multi-area RNN also provides several testable hypotheses. First, because the IB simplifies the readout, it suggests a simple readout from “output” areas of cortex (e.g PMd or Motor Cortex). In our multi-area RNN, we found that the output area was comprised of pools of neurons that represent left or right reaches29, enabling a simple readout (Fig. S13b, c.). In particular, the connectivity matrix for our area 3 was composed of two pools of excitatory neurons with a common inhibitory pool. Overall, this connectivity and winner-take-all architecture is consistent with our PMd activity, which appears to implement winner-take-all dynamics between a pool of neurons representing left and right reaches (Fig. S13h,i.) Second, due to selective propagation of inter-areal information, direction axis activity in DLPFC should be more predictive of activity in downstream regions such as PMdr and PMd than activity in the top PCs. Simultaneous recordings of DLPFC and PMd would help test this hypothesis.

Materials and methods

Task and training details

Somatomotor reaction time visual discrimination task and recordings from DLPFC and PMd

The task, training and electrophysiological methods used to collect the data used here have been described previously 23 and are reviewed briefly below. All surgical and animal care procedures were performed in accordance with National Institutes of Health guidelines and were approved by the Stanford University Institutional Animal Care and Use Committee and the Boston University Institutional Animal Care and Use Committees.

Two trained monkeys (Ti and Ol) performed a visual reaction time discrimination task. The monkeys were trained to discriminate the dominant color in a central static checkerboard composed of red and green squares and report their decision with an arm movement. If the monkey correctly reached to and touched the target that matched the dominant color in the checkerboard, they were rewarded with a drop of juice. This task is a reaction time task, so that monkeys initiated their action as soon as they felt they had sufficient evidence to make a decision. On a trial-by-trial basis, we varied the signed color coherence of the checkerboard, defined as (R − G)/(R + G), where R is the number of red squares and G the number of green squares. The color coherence value for each trial was chosen uniformly at random from 14 different values arranged symmetrically from 90% red to 90% green. Reach targets were located to the left and right of the checkerboard. The target configuration (left red, right green; or left green, right red) was randomly selected on each trial. Both monkeys demonstrated qualitatively similar psychometric and reaction-time behavior. 996 units were selected from PMd of Ti (n=546) and Ol (n=450) and 2819 units were recorded from DLPFC of Ti while they performed the task23. Monkey Ol and Ti’s PMd units both had low choice color probability. Reported analyses from PMd data use units pooled across Monkey Ol and Ti.

RNN description and training

We trained a continuous-time RNN to perform the checkerboard task. The RNN is composed of N artificial neurons (or units) that receive input from Nin time-varying inputs u(t) and produce Nout time-varying outputs z(t). The RNN defines a network state, denoted by x(t) ∈ ℝN; the ith element of x(t) is a scalar describing the “currents” of the ith artificial neuron. The network state is transformed into the artificial neuron firing rates (or network rates) through the transformation:

where f (·) is an activation function applied elementwise to x(t). The activation function is typically nonlinear, endowing the RNN with nonlinear dynamics and expressive modeling capacity 46. In this work, we use f (x) = max(x, 0), also known as the rectified linear unit, i.e., f (x) = relu(x). In the absence of noise, the continuous time RNN is described by the equation

where τ is a time-constant of the network, Wrec ∈ ℝN×N defines how the artificial neurons are recurrently connected, brec ∈ ℝN defines a constant bias, maps the RNN’s inputs onto each artificial neuron, and Et is the recurrent noise. The output of the network is given by a linear readout of the network rates, i.e.,

where maps the network rates onto the network outputs.

We trained RNNs to perform the checkerboard task as follows. For all networks, unless we explicitly varied the amount of units, we used Nin = 4, N = 300, and Nout = 2.

The four inputs were defined as:

Whether the left target is red (−1) or green (+1).

Whether the right target is red (−1) or green (+1).

Signed coherence of red (ranging from −1 to 1), (R − G)/(R + G).

Signed coherence of green (ranging from −1 to 1), (G − R)/(R + G). Note that, prior to the addition of noise, the sum of the signed coherence of red and green is zero.

The inputs, u(t) ∈ ℝ4, were defined at each time step, t, in distinct epochs. In the ‘Center Hold’ epoch, which lasted for a time drawn from distribution 𝒩 (200 ms, 502 ms2), all inputs were set to zero. Subsequently, during the ‘Targets’ epoch, which lasted for a time drawn from distribution 𝒰[600 ms, 1000 ms], the colors of the left and right target were input to the network. These inputs were noiseless, as illustrated in Fig. 3a, to reflect that target information is typically unambiguous in our experiment. Following the ‘Targets’ epoch, the signed red and green coherences were input into the network during the ‘Decision’ epoch. This epoch lasted for 1500 ms. We added zero mean independent Gaussian noise to these inputs, with standard deviation equal to 5% of the range of the input, i.e., the noise was drawn from 𝒩 (0, 0.12). At every time point, we drew independent noise samples and added the noise to the signed red and green coherence inputs. We added recurrent noise Et, adding noise to each recurrent unit at every time point, from a distribution 𝒩 (0, 0.052). Following the ‘Decision’ epoch, there was a ‘Stimulus Off’ epoch, where the inputs were all turned to 0.

The two outputs, z(t) ∈ ℝ2 were defined as:

Decision variable for a left reach.

Decision variable for a right reach.

We defined a desired output, zdes(t), which was 0 in the ‘Center Hold’ and ‘Targets’ epochs. During the ‘Decision’ epoch, zdes(t) = 1. In the ‘Stimulus Off’ epoch, zdes(t) = 0. In RNN training, we penalized output reconstruction using a mean-squared error loss,

The set 𝒯 included all times from all epochs except for the first 200 ms of the ‘Decision’ epoch from the loss. We excluded this time to avoid penalizing the output for not immediately changing its value (i.e., stepping from 0 to 1) in the ‘Decision’ epoch. Decision variables are believed to reflect a gradual process consistent with non-instantaneous integration of evidence, e.g., as in drift-diffusion style models, rather than one that steps immediately to a given output.

To train the RNN, we minimized the loss function:

where

||A||F denotes the Frobenius norm of matrix A.

λin = λrec = λout = 1, λr = 0 to penalize larger weights.

λΩ = 2.

ℒΩ is a regularization term that ameliorates vanishing gradients proposed and is described in prior literature26,47.

During the training process, we also incorporated gradient clipping to prevent exploding gradients47. Training was performed using stochastic gradient descent, with gradients calculated using backpropagation through time. For gradient descent, we used the Adam optimizer, which is a first order optimizer incorporating adaptive gradients and momentum48.

Every 200 or 500 training epochs, we generated 2800 cross-validation trials, 100 for each of the 28 possible conditions (14 coherences × 2 target configurations). For each trial, there was a correct response (left or right) based on the target configuration and checkerboard coherence. When training, we defined a “correct decision” to be when the RNNs DV for the correct response was greater than the other DV and the larger DV was greater than a pre-set threshold of 0.6. We evaluated the network 500ms before the checkerboard was turned off (the end of the trial). We required this criteria to be satisfied for at least 65% of both leftward and rightward trials. We note that this only affected how we terminated training. It had no effect on the backpropagated gradients, which depended on the mean-squared-error loss function. Note that a trial that outputted the correct target but did not reach the 0.6 threshold would not be counted towards the 65% criteria.

When testing, we defined the RNNs decision to be either: (1) whichever DV output (for left or right) first crossed a pre-set threshold of 0.6, or (2) if no DV output crossed the pre-set threshold of 0.6 by the end of the ‘Decision epoch,’ then the decision was for whichever DV had a higher value at the end of this epoch — an approach that is well established in models of decision-making 49,50. If the RNN’s decision on a single trial was the same as the correct response, we labeled this trial ‘correct.’ Otherwise, it was incorrect. The proportion of decisions determined under criterion (2) was negligible (0.5% across 100 trials for each of 28 conditions). An interpretation for criterion (2) is that if the RNN’s DV has not achieved the threshold certainty level by the end of a trial, we assign the RNN’s decision to be the direction for which its DV had the largest value. Finally, in training only, we introduced ‘catch’ trials 10% of the time. On 50% of catch trials, no inputs were shown to the RNN and zdes(t) = 0 for all t. On the remaining 50% of catch trials, the targets were shown to the RNN, but no coherence information was shown; likewise, zdes(t) = 0 for all t on these catch trials.

We trained the three-area RNNs by constraining the recurrent weight matrix Wrec to have connections between the first and second areas and the second and third areas. In a multi-area network with N neurons and m areas, each area had N/m neurons. In our 3-area networks, each area had 100 units. Of these 100 units, 80 were excitatory and 20 were inhibitory. Excitatory units were constrained to have only positive outgoing weights, while inhibitory units were constrained to have only negative outgoing weights. We used the pycog repository 26 to implement these architecture constraints. The parameters for the exemplar RNN used in the paper are shown in Table 1. In our hyperparameter sweeps, we varied the hyperparameters of the exemplar RNN. For each parameter configuration, we trained 8 different networks with different random number generator seeds. For analyses of the RNN, we fixed the timing of trials, obviating the need to to restretch trial lengths. Note that while at inference, we generated RNN trials with equal length, the RNN was trained with varying delay periods.

Additional description of analyses

Decoding analysis for DLPFC and PMd data

For DLPFC and PMd data, we calculated decoding accuracy using 400 ms bins. We report numbers in a window [−300ms, +100 ms] aligned to movement onset. We used the Python sklearn.svm.SVC command with 80% training and 20 % test sets. Decoding analyses were performed on single trials using 2-49 simultaneously recorded units from Plexon U-probes and the averages reported are across PMd and DLPFC sessions, respectively. To assess whether decoding accuracies were significant, we shuffled the trials for 100 times. The decoding accuracy for direction, color and target configuration variables of a session was judged to be significant if it lies above the 99 percentile of shuffled accuracy. For DLPFC, direction, target configuration and color decoding accuracy of 100%, 78%, 53% sessions were judged to be significantly above chance; for PMd, 97%, 26%, 20% sessions demonstrated significant decoding accuracy to direction, target configuration and color.

Mutual information was calculated by computing H(Y) − LCE where H(Y) was 1 and LCE denotes the cross entropy loss (in bits). We computed the decoding and information only for sessions with decoding accuracy significantly above chance. Negative mutual information was set to zero.

Decoding and Mutual information for RNNs

We used a decoder and mutual information approximation to quantify the amount of information (color, target configuration, direction) present in the network. In particular, we trained a neural network to predict the reported direction choice, color choice, and target configuration from the single trial activity of a population of units. We focused on decoding the network’s choice (as described above in “RNN description and training” of Methods) rather than the ground-truth input information to focus on how the behavioral choice was represented across areas. We note that the network’s choice and input information are highly correlated in trained networks (Fig. 1b and Fig. S1). We used 700 trials for training, and 2100 independent trials for testing. To generate the trials for training and testing, we increased the recurrent noise to be drawn from the distribution (𝒩 (0, 0.12)) to prevent overfitting. For each trial, we averaged data in a window [−300ms, +100ms] around reaction time.

We trained a neural network with 3 layers, 64 units per layer, leakyRelu activation (α=0.2), and dropout (p=0.5), using stochastic gradient descent, to predict the choice given the activity of the population. We trained the neural network to minimize the cross-entropy loss. We used the same neural network from the decode to compute an approximation to mutual information, described in Supplementary Note 2. In previous work29, we obtained analogous results using a linear decoder across hyperparameter sweeps. We also obtain analogous results by using an SVM decoder in Fig. S4 for our exemplar parameter configuration.

RNN behavior

To evaluate the RNN’s psychometric curve and reaction-time behavior, we generated 200 trials for each of the 28 conditions, producing 400 trials for each signed coherence. For these trials, we calculated the proportion of red decisions by the RNN. This corresponds to all trials where the DV output for the red target first crossed the preset threshold of 0.6; or, if no DV output crossed the threshold of 0.6, if the DV corresponding to the red target exceeded that corresponding to the green target. The reaction time was defined to be the time between checkerboard onset to the first time a DV output exceeded the preset threshold of 0.6. If the DV output never exceeded a threshold of 0.6, in the reported results, we did not calculate a RT for this trial.

dPCA

Demixed principal components analysis (dPCA) is a dimensionality reduction technique that provides a projection of the data onto task related dimensions while preserving overall variance51. dPCA achieves these aims by minimizing a loss function:

Here, Xc refers to data averaged over a “dPCA condition” (such as time, coherence, target configuration, color, or direction), having the same shape as X ∈ ℝN×cT, but with the entries replaced with the condition-averaged response. The aim is to recover (per dPCA condition c) a Pc and Dc matrix. Pc is constrained to have orthonormal columns, while Dc is unconstrained. The number of columns of Pc and rows of Dc reflects the number of components one seeks to find per condition. The column of Pc reflects how much the demixed data contributes to each neuron. We use the principal axes from Pc to compute the axis overlap, as in Kobak et al51. We used axes of dimension 1 for RNNs, which were sufficient to capture most color, target configuration, or direction variance. For the neural data, we used five components for direction, color and target configuration since the PMd data was higher dimensional than the RNNs.

For multi-area analyses, we separated the units for each area and found the task-relevant axes for this subset of units. For the inter-area analyses, we used RNNs with only excitatory connections, and therefore found the color and direction axis using only the excitatory units. In all other analyses, all units were used to identify the axes. For RNN activity, we performed dPCA using activity over the entire trial. For analyses of the RNN, we fixed the timing of trials, obviating the need to to restretch trial lengths. Note that while at inference, we generated RNN trials with equal length, the RNN was trained with varying delay periods.

For neural data, due to the stochasticity in task design, there is a trial-by-trial difference in interval between target and checkerboard onset (TC interval). The reaction time (from the checkerboard onset to monkey’s hand movement initiation) also varies for each trial. To align time events across trials, we restretched the firing rates in each trial. For DLPFC units, each trial was aligned to targets onset first. Median reaction time (527ms) and TC interval (735ms) were calculated by combining every trial in the database. For each trial, TC interval and reaction time was either compressed or stretched to the median values through linear interpolation. After the data restretching, we choose the data window T as 1300ms, from −100ms to 1200ms around target onset with sample size of 1ms. For every unit n in total units number N, we averaged the single-trial firing rate by stimulus S (checkerboard dominant color, green or red) and decision of choice D (left or right). As a result, a 4D firing-rate matrix XN×S×D×T was created as input to demixed principal component analysis algorithm.

For PMd units, activities before checkerboard onset were minimal. As a result, each trial was aligned to target onset and a segment with time window of [−100ms, 367ms] was chosen first. Then the same trial was aligned to checkerboard first and a segment with a window of [−368ms, 465ms] was chosen. The final restretched data was the concatenation of these two data segments.

When computing the overlap in Fig. 4, we averaged across 8 initializations, and computed the PSTHs over 700 trials. In our dpca variance sweeps (Fig. 5), we computed the PSTHs over 280 trials.

PCA

Principal components analysis (PCA) is a dimensionality reduction technique that projects high-dimension data into low-dimensional axis which maximize the variance in the data. PCA provides low-dimensional projections by minimizing the loss function:

XN×T is high-dimension raw data and DN×N is the decoding matrix. The low-dimension trajectories xM×T (M < N) calculated by multiplying first M rows of D by X.

Before applying PCA on the data, the raw data was preprocessed by data normalization and average firing rate removal:

Data normalization for PCA for neural data

Due to the hererogenity in firing rates, PCA results might be dominanted by units with high firing rates. To equalize the contribution of each unit in PCA, for each unit n ∈ 𝒩, 99 percentile of max firing rate xn,99 was calculated over all stimulus, decision conditions and time point. Then we divided each data for unit xn,99 by square root of to normalize the data.

Condition independent signal removal for PCA

The condition independent signal is another source that explains substantial amount of population variance other than task-related signal. Before conducting principal component analysis (PCA), we calculated the average firing rate of each single unit X1×1×1×T over all stimulus and decision conditions and subtracted this condition independent signal from the time-restretched data X1×S×D×T. We also visualized the PCs of the RNNs in an analogous manner, unless otherwise noted.

Canonical correlation

We applied CCA to assess the similarity between neural activity and the artificial unit activity 52. Before applying CCA, we performed principal component analysis to reduce the dimensionality of the artificial and neural activity to remove noise, which can be arbitrarily reshaped to increase the canonical correlation 52. We reduced the dimensionality of PMd data to 2 (which captures over 80% of the PMd variance). For DLPFC, 18 dimensions were required to capture over 80% of the variance, but at such a high dimensionality, noise can be reshaped to significantly increase the canonical correlations. For DLPFC, we therefore show a comparison to the top 4 PCs in Fig. 3e. However, the trends held irrespective of the DLPFC dimensionality we chose, as shown in Fig. S3. We report the average CCA correlation coefficients in Fig. 3e using times in a window of [−400ms, 400ms] aligned to checkerboard onset. The data was binned in 10ms bins.

Analyses of inputs and activity

In order to disentangle the effects of external inputs and recurrence, in Fig. S9, we evaluated the input contribution and overall activity. For Area 1, we defined the input contribution as Winut, and for areas 2 and 3, we defined the input contribution as , and respectively, where denotes the activity of the units in area m. The activity corresponds to the firing rate that experimentalists could measure, reflecting a combination of input and recurrent interactions. For constant inputs, a stable value of the activity implies there is little recurrent processing.

Inter-Area Projection Analyses

To calculate the overlap between the color and direction axes with the potent and null spaces, we performed singular value decomposition on the inter-area connections, W21 and W32. W21 and W32 were 80 × 80 matrices, and were full rank. Nevertheless, they had near some zero singular values, indicating that the effective rank of the matrix was less than 80. We defined the potent dimensions to be the top m right singular vectors, while the null dimensions were the remaining 80 − m right singular vectors.

We performed the analyses of Fig. 4f by varying the potent and null dimensions, sweeping m from 1 to 80. For each defined potent and null space, we calculated the axis overlap between the direction (or color) axis and the potent (or null) space by computing the L2-norm of the orthogonal projection (squared). We report the squared quantity because the expectation of the norm of a projection of a random vector onto an m-dimensional subspace of an n-dimensional space is m/n. We include an approximation of the expectation of the projection of a random vector in Fig. 4 by averaging the projection of 100 random vectors. Our results show that the direction axis was always more aligned with potent dimensions than the color axis, irrespective of the choice of m, and that the direction axis was preferentially aligned with the top singular vector.